
<h3>Installing required software</h3>
The only prerequisite for the R training is to install the latest version of R and RStudio on your computer and the packages we're using for the training. These are available online and are free to download and install. We'll talk about the difference between R and RStudio on the first day, but for now, just make sure they're installed. 

<ol>
<li><b>Install R:</b> Download and install R 4.5.3 <a href="https://cran.r-project.org/bin/windows/base/">for Windows</a> or for <a href="https://cran.r-project.org/bin/macosx/big-sur-arm64/base/R-4.5.3-arm64.pkg"> for macOS</a>. 
  <ul>
  <li>If you run into issues with the link for Windows, go back to the <a href="https://cran.r-project.org/bin/windows/">Windows install page</a>, and click on <a href="https://cran.r-project.org/bin/windows/base/">install for the first time</a>, then <a href="https://cran.r-project.org/bin/windows/base/R-4.5.3-win.exe">Download R-4.5.3 for Windows</a> link.</li>
  <li>If you have issues with the macOS link, try downloading R-4.5.3-arm64.pkg from the <a href="https://cran.r-project.org/bin/macosx/">macOS install page</a>.</li>
  </ul><br>
<li><b>Install R Studio:</b> The main <a href="https://posit.co/download/rstudio-desktop/">RStudio page</a> has links to download <a href = "https://download1.rstudio.org/electron/windows/RStudio-2026.01.1-403.exe">RStudio-2026.01.01-403.EXE</a> for Windows (assuming your operating system is at least Windows 10), and <a href='https://download1.rstudio.org/electron/macos/RStudio-2026.01.1-403.dmg'>RStudio-2026.01.01-403.DMG</a> for macOS (assuming OS 14 or greater)</li>
</li>
</ol>

<p class = 'note'>
<b>NOTE:</b> If you already have R and RStudio on your computer, the R version should be at least 4.4.3 and the RStudio version at least 2025.06.2-418 to make sure everyone's code behaves the same way. You can check your software versions by opening RStudio. The console in the bottom left will display the R version. Check the RStudio version by going to Help > About RStudio. The first 4 numbers should be 2025 or higher.</p>

</br>
<details open><summary class='drop'><b style="color:red">Required Packages</b></summary>
A number of packages are required to follow along with data wrangling and visualization sessions. Please try to install these in RStudio ahead of time by running the code below. If you don't know how to run the code, open view the <b>Running Code Screencast</b> below for how to do this. 
```{r reqpkg, echo = T, eval = F}
packages <- c("tidyverse", # for Day 2 and 3 data wrangling
              "RColorBrewer", "viridis", "patchwork", # for Day 3 ggplot
              "readxl", "writexl") # for day 1 importing from excel

install.packages(setdiff(packages, rownames(installed.packages())))  

# Check that installation worked
library(tidyverse) # turns on core tidyverse packages
library(RColorBrewer) # palette generator
library(viridis) # more palettes
library(patchwork) # multipanel plots
library(readxl) # reading xlsx
library(writexl) # writing xlsx
```

Note that when you turn on the tidyverse package via `library(tidyverse)`, you will get the following message in your console. This means the tidyverse successfully loaded on your machine and is not an error. 
<center>
<img src = "./images/tidyverse_not_error.png" alt = 'tidyverse message' style = "display:block;width:750px;">
</center>
<br>

<details><summary class='drop2'>Running Code Screencast</summary>
<video height = "650" controls>
<source src="images/Installing_packages.mp4" type="video/mp4">
</video>
</details>
</details>
<br>

<details open><summary class = 'drop'><b style="color:red">Check you can download data from github</b></summary>
We'll download these as we run through the code. The check is that your computer doesn't have security restrictions that prevent importing datasets from github into R. All datasets used in this training will also be available via a zip file. The github hosted datasets are just easier to make sure everyone downloads the same version and is grabbing them from the same place. 

<p class = 'code'>Copy the code below into RStudio and run it. If you don't get an error message, you're good.</p> If you do get an error message, you'll just need to import the datasets locally using the provided zip file. 
```{r datasets, eval = F}
motinv <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv")
```

</details>

<details open><summary class='drop'>Optional Reading</summary>
This is completely optional, but if you have any time before training starts, I highly recommend reading <a href="https://stat545.com/r-basics.html">Chapter 2: R basics and workflows</a> in STAT545. This is an online book based on a graduate level statistics class that Dr. Jenny Bryan teaches and is turning into this book. She's a stats professor turned senior programmer at RStudio, and I really like how she approaches programming R in this book. 
</details>
<br>

<details open><summary class='drop'>About the Training</summary>
<ul>
<li><b>Timing:</b> The training will take place over three consecutive Fridays (4/3, 4/10, 4/17). Each day will run from 12:30 - 3:30pm EST. We'll take short breaks between topics. </li>

<li><b>Office Hours:</b> For the day prior to training, I will be available virtually from 12:30 - 2pm EST for office hours in case there are questions that couldn't be handled before or during training. I will send a Teams link to participants for this office hours via email. Folks also welcome to email me questions.</li>

<li><b>Structure:</b> For most of the training, I will share my screen as I go through the website and then demo with live coding. Participants should have their own laptop to follow along with the code. </li>

<li><b>Objectives:</b> Three days is barely enough to scratch the surface on what you can do in R. My goals with this training are to:
<ol>
<li>Help you get beyond the initial learning curve that can be really challenging to climb on your own.</li>
<li>Expose you to what I consider are the useful things R can do for us. </li>
<li>Provide you the tools needed to continue advancing your R skills on your own.</li>
</ol>
</li>

<li><b>Credit:</b> Much of this training was borrowed heavily from <a href = "https://katemmiller.github.io/IMD_R_Training_Intro/">NPS IMD Intro to R</a> training in 2022. A ton of credit for this training goes to the developers of those lessons:
<ul>
<li>Day 1 Intro to R: Sarah Wright and Andrew Birch</li>
<li>Day 2 Data Wrangling: John Paul Schmit and Lauren Pandori</li>
<li>Day 3 Data Visualization: Ellen Cheng and Kate Miller (Spatial Data) </li>
<li>Day 4 Programming Best Practices: Sarah Wright and Thomas Parr </li>
</ul>
</li>

<li><b>Feedback:</b> To help me improve this training for future sessions, please leave feedback in the training feedback form in the initial email. You can submit feedback multiple times and don't need to answer every question. Responses are anonymous.</li>
</ul>
</details>
<br>

