
#### Basic Plotting
<details open><summary class = 'drop'>Basic plotting</summary>
Visualizing the data is also important to get a sense for the data and look for potential errors and outliers. Base R has plotting functions that allow you to create quick plots without having to know a lot of code. I often use Base R plot functions when I'm exploring data but not making plots I plan to use for publication. When I need to create more complex plots, I use `ggplot2`, which we'll cover on Day 2 and 3.

Histograms are a great start. The code below generates a basic histogram plot of a specific column in the dataframe using the `hist()` function.

```{r day1_bp, echo = F, include = F}
#------------------------------------
#     Day 1: Basic Plotting Code 
#------------------------------------
```

<p class='code'>Plot histogram of motile invertebrate No.Damage counts </p>
```{r hist, fold.plot = F, fig.height = 4, fig.width = 6}
hist(x = motinv3$No.Damage)
```
<br>
Looking at the histogram, it looks like all of the counts are below 500cm except for one that's way out in 2000 range. You can also make a scatterplot of the data. If you only specify one column, the x axis will be the row number for each record, and the y axis will be the specified column.

<p class='code'>Make point plot of No.Damage counts</p>
```{r pdbh1, fold.plot = F, fig.height = 4, fig.width = 6}
plot(motinv3$No.Damage)
```
<br>
Again, you can see there's one value that's greater than all of the others. 

We can also plot two variables in a scatterplot. 

<p class='code'>Make scatterplot of No.Damage vs. Damage_num (Option 1)</p>
```{r pdbh2, fold.plot = F, fig.height = 4, fig.width = 6}
plot(motinv3$No.Damage ~ motinv3$Damage_num)
```

<p class='code'>Make scatterplot of No.Damage vs. Damage_num (Option 2- better axis labels)</p>
```{r pdbh5, fold.plot = F, fig.height = 4, fig.width = 6}
plot(No.Damage ~ Damage_num, data = motinv3) # equivalent but cleaner axis titles
```
<br>
Here you can see there's one value that's greater than all of the others in both sets of counts. These would be worth looking at more carefully to determine if they're errors in the data. 

<div class="alert alert-info">
<strong>CHALLENGE: Plot a histogram of `Damage_num` in the motinv3 data frame</strong>

<details><summary class='code2'>Answer</summary>
```{r wethist, fig.height = 4, fig.width = 6, fold.plot = F}
hist(motinv3$Damage_num)
```
</details>
</div>
</details>
<br>

