#### Day 1 Questions {.tabset}
<details open><summary class = 'drop'>Load Data </summary>
If you're starting a new R session to answer these questions, you'll need to read in the wetland and tree data frames again.

```{r day1_ch, echo = F, include = F}
#------------------------------------
#     Day 1: Challenges Code 
#------------------------------------
```

<p class = 'code'>Read in example Bass Harbor motile invertebrate data from url</p>
```{r c1rdatc1, eval = T}
motinv <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv"
  )
```

<p class = 'code'>Read in example NETN tree data from url</p>
```{r c1readtree, eval = F}
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
```
</details><br>

<details open><summary class = 'drop'>Data Structures</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you look at the the first 4 even rows (2, 4, 6, 8), and first 2 columns of the `motinv` data frame?</strong>

<details><summary class='drop2'>Answer</summary>
<h5>Answer that works<h5>
```{r chal1d1c, fold.output = F}
motinv[c(2, 4, 6, 8), c(1, 2)]
```

<h5>Better answer that's more stable</h5>
```{r chal1bd1c, fold.output = F}
names(motinv) # get the names of the first 2 columns
motinv[c(2, 4, 6, 8), c("Network", "UnitCode")]
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How many unique species are there in the `motinv` data frame?</strong>

<details><summary class='drop2'>Answer</summary>
```{r rows2ad1c, results = "hide"}
# Option 1
length(unique(motinv[, "ScientificName"])) # 6
```

```{r rows2d1cb, fold.output = FALSE}
# Option 2
length(unique(motinv$ScientificName)) # equivalent
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What years were QAQC visits conducted (QAQC = TRUE)?</strong>
<details><summary class='drop2'>Answer</summary>
```{r, rows2a2d1c, results = 'hide'}
# Option 1 - used unique to just return unique site name
unique(motinv$Year[motinv$QAQC == TRUE]) # 2013
```

```{r, rows2b2d1c, fold.output=FALSE}
# Option 2
unique(motinv[motinv$QAQC == TRUE, "Year"])
```
</details>
</div>
</details><br>

<details open><summary class = 'drop'>Data Exploration</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using motinv3, how many species are found in PlotName A1 in 2024?</strong>
<details><summary class='drop2'>Answer</summary>
Option 1. Use brackets, then calculate the number of rows.
```{r chal4c1}
# with brackets
A1_2024 <- motinv3[motinv3$PlotName == "A1" & motinv3$Year == 2024, ]
nrow(A1_2024) # 3
```
Option 2. Use base R subset, then view the data.frame to see.
```{r chal4bc1, eval = F}
# with base R subset
A1_2024b <- subset(motinv3, PlotName == "A1" & Year == 2024)
View(A1_2024b) # 3
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What years have green crabs (Latin: Carcinus maenas) been detected? </strong>
<details><summary class='drop2'>Answer</summary>

Option 1. Subset the data with brackets and use the `sort(unique())` to give an easier to read output.
```{r chal5bc1, fold.output = F}
# OPTION 2
gcrab <- motinv3[motinv3$Species == "Carcinus maenas",]
sort(unique(gcrab$Year)) #2019, 2021, 2022, 2023, 2024

```

Option 2. Subset data then use `table()` to tally the years and number of rows green crabs were found.
```{r chal5c1, fold.output = F}
gcrab2 <- subset(motinv3, Species == "Carcinus maenas")
table(gcrab2$Year)
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find the highest value recorded in No.Damage column.</strong>
<details><summary class='drop2'>Answer</summary>
There are multiple ways to do this. Two examples are below.

Option 1. View the data and sort by No.Damage.
```{r vmotc1, eval = F}
View(motinv3)
```

Option 2. Find the max No.Damage count and subset the data frame
```{r maxndc1, fold.output = F}
max_nd <- max(motinv3$No.Damage, na.rm = TRUE)
motinv3[motinv3$No.Damage == max_nd,]
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Fix the No.Damage typo by replacing 1960 with 196.</strong>

<details><summary class = 'drop2'>Answer</summary>
Let's say that you looked at the datasheet, and the actual count for No.Damage was 196 instead of 1960. You can change that value in the original CSV by hand. But even better is to document that change in code. There are multiple ways to do this. Two examples are below.

But first, it's good to create a new data frame when modifying the original data frame, so you can refer back to the original if needed. I also use a really specific filter to make sure I'm not accidentally changing other data.

<p class = 'code'>Replace 1960 with 196</p>
```{r trfxc1, fold.output = F}
# create copy of motinv data
motinv_fix <- motinv3

# find the problematic value, and change it to 196
motinv_fix$No.Damage[motinv_fix$Year == 2019 & 
                       motinv_fix$PlotName == "R4" & 
                       motinv_fix$No.Damage == 1960] <- 196

# check your work
range(motinv3$No.Damage) #1960
range(motinv_fix$No.Damage) # now 282

```
</details>
</div>
</details><br>

<details open><summary class = 'drop'>Basic Plotting</summary>
<div class="alert alert-info">
<strong>Plot a histogram of percent cover (`Ave_Cov`) in the `ACAD_wetland` </strong>

<details><summary class='code2'>Answer</summary>
```{r wethist2, eval = F}
hist(ACAD_wetland$Ave_Cov)
```
</details>
</div>
</details>
<br>