#### Day 2 Questions
<details open><summary class = 'drop'>Load Data and Packages</summary>
If you're starting a new session to answer these questions, you'll need to load dplyr and read in the motile invertebrate data frame again.<br>

<p class = 'code'>Load dplyr</p>
```{r day2_ch, echo = F, include = F}
#---- Day 2: Challenges Code ----
```

```{r ldpd2c}
library(dplyr)
```

<p class = 'code'>Read in example motile invertebrate and point intercept data</p>
```{r readmotinv3bc1, eval = F}
motinv <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv"
  )

pi_dat <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_Point_Intercept_data.csv"
)
```

```{r motinvd2c1, echo = F}
motinv <- read.csv("./data/BASHAR_motile_invert_counts.csv")
pi_dat <- read.csv("./data/BASHAR_Point_Intercept_data.csv")
```

</details>
<br>

<details open><summary class = 'drop'>Data Wrangling with dplyr </summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using `motinv`, how many species are found in PlotName = A1 in 2024?</strong>
<details><summary class='drop2'>Answer</summary>

```{r chal4dpc1}
# with brackets
A1_2024 <- motinv |> filter(PlotName == "A1" & Year == 2024)
nrow(A1_2024) # 3
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What years have green crabs (Latin: Carcinus maenas) been detected? </strong>
<details><summary class='drop2'>Answer</summary>

```{r chal5bdpc1, fold.output = F}
gcrab <- motinv |> filter(ScientificName == "Carcinus maenas") |> 
  select(Year) |> unique()

gcrab
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find the highest value recorded in No.Damage column.</strong>
<details><summary class='drop2'>Answer</summary>
```{r maxnddpc1, fold.output = F}
max_nd <- max(motinv$No.Damage, na.rm = TRUE)
motinv3 |> filter(No.Damage == max_nd)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Fix the No.Damage typo by replacing 1960 with 196.</strong>

<details><summary class = 'drop2'>Answer</summary>
Let's say that you looked at the datasheet, and the actual count for No.Damage was 196 instead of 1960. You can change that value in the original CSV by hand. But even better is to document that change in code. There are multiple ways to do this. Two examples are below.

But first, it's good to create a new data frame when modifying the original data frame, so you can refer back to the original if needed. I also use a really specific filter to make sure I'm not accidentally changing other data.

<p class = 'code'>Replace 1960 with 196</p>
```{r trfxc1d2, fold.output = F}
# dplyr approach
motinv_fix <- motinv |> mutate(No.Damage = replace(No.Damage, No.Damage == 1960, 196))
range(motinv$No.Damage)
range(motinv_fix$No.Damage)

```
</details>
</div>
</details>

<br>

<details open><summary class = 'drop'>Conditionals</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using the motinv data, create a new column called trophic that indicates whether the species is an herbivore or predator. Hint: predator site codes are c("CARMAE", "NUCLAP"), and herbivore site codes are c("LITLIT", "LITOBT", "LITSAX", "TECTES"). </strong>
<details><summary class='code2'>Answer</summary>
```{r micondc1}
herb <- c("LITLIT", "LITOBT", "LITSAX", "TECTES")
pred <- c("CARMAE", "NUCLAP")

# base R
motinv$trophic <- ifelse(motinv$SpeciesCode %in% pred, "predator", "herbivore")
table(motinv$trophic, motinv$SpeciesCode)

# tidyverse
motinv <- motinv |> mutate(trophic = ifelse(SpeciesCode %in% pred, "predator", "herbivore"))
table(motinv$trophic, motinv$SpeciesCode)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the motile invertebrate data, create a new column called count_level that has levels High, Medium, Low, based on No.Damage, where "High" is > 35, "Medium" is 10 - 35, and "Low" is < 10. </strong>
<details><summary class='code2'>Answer</summary>
```{r miabc1}
# Base R using the with() function and nested ifelse()
motinv$count_level <- 
  ifelse(motinv$No.Damage > 35, "High", 
         ifelse(motinv$No.Damage >=10 & motinv$No.Damage <=35, "Medium", "Low"))
table(motinv$count_level)
```

```{r miab2c1}
# Tidyverse using case_when() and between
motinv <- motinv |> mutate(count_level = case_when(No.Damage > 35 ~ "High",
                                                   between(No.Damage, 10, 35) ~ "Medium",
                                                   No.Damage < 10 ~ "Low"))
table(motinv$count_level)
```
Note the use of the `between()` function that saves typing. This function matches as >= and <=. 
</details>
<br>

<details open><summary class='drop'>Summarizing</summary>
<div class="alert alert-info">
<strong>Using the ACAD_wetland data, calculate the average percent cover of native vs. invasive species per plot. Note that Invasive = TRUE is invasive and FALSE is native. </strong>
<details><summary class='code2'>Answer</summary>
```{r minvcovc1, eval = F}
# Using group_by()
ACAD_inv <- ACAD_wetland |> group_by(Site_Name, Year, Invasive) |> 
  summarize(Pct_Cov = sum(Ave_Cov), 
            .groups = 'drop') |>  # optional line to keep console from being chatty
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_inv)

# Using summarize(.by)
ACAD_inv2 <- ACAD_wetland |> 
  summarize(Pct_Cov = sum(Ave_Cov), .by = c(Site_Name, Year, Invasive)) |> 
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_inv2) # should be the same as ACAD_inv
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the ACAD_wetland data, count the number of native vs. invasive species per plot. Note that Invasive = TRUE is invasive and FALSE is native. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetinvrc1, eval = F}
# Using group_by()
ACAD_spp <- ACAD_wetland |> group_by(Site_Name, Year, Invasive) |> 
  summarize(num_spp = n(), 
            .groups = 'drop') |>  # optional line to keep console from being chatty
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_spp)

# Using summarize(.by)
ACAD_spp2 <- ACAD_wetland |> 
  summarize(num_spp = n(), .by = c(Site_Name, Year, Invasive)) |> 
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_spp2) # should be the same as ACAD_inv
```
</details>
</div>

<div class="alert alert-info">
<strong>Using ACAD_wetland data, calculate relative % cover of species within each site. This one is challenging! </strong>
<details><summary class='code2'>Answer</summary>
<p class = 'code'>First sum site-level cover using mutate to return a value for every original row.</p>
```{r wetrelcovc1, eval = F}
ACAD_wetland <- ACAD_wetland |> group_by(Site_Name, Year) |> 
  mutate(Site_Cover = sum(Ave_Cov)) |> 
  ungroup() # good practice to ungroup after group.

table(ACAD_wetland$Site_Name, ACAD_wetland$Site_Cover) # check that each site has a unique value.
head(ACAD_wetland)
```

<p class = 'code'>Next calculate relative cover grouped on Site_Name, Year, Latin_Name, and Common </p> 
```{r wetrelcov2c1, eval = F}
# Create new dataset because collapsing rows on grouping variables
# Using group_by() and summarize()
ACAD_wetland_relcov <- ACAD_wetland |> group_by(Site_Name, Year, Latin_Name, Common) |> 
  summarize(rel_cov = (Ave_Cov/Site_Cover)*100,
            .groups = 'drop') |> 
  ungroup()

```

<p class = 'code'>Check that relative cover sums to 100% within each site</p>
```{r wetrelcovcheckc1, eval = F}
# Using summarize(.by = ) 
ACAD_wetland_relcov2 <- ACAD_wetland |> #group_by(Site_Name, Year, Latin_Name, Common) |> 
  summarize(rel_cov = Ave_Cov/Site_Cover, 
            .by = c("Site_Name", "Year", "Latin_Name", "Common"))

# Check that your relative cover sums to 100 for each site
relcov_check <- ACAD_wetland_relcov2 |> group_by(Site_Name, Year) |> 
  summarize(tot_relcov = sum(rel_cov)*100, .groups = 'drop')
table(relcov_check$tot_relcov) # they should all be 100

```
</details>
</div>
</details>
<br>
<details open><summary class = 'drop'>Plotting with ggplot2</summary>
<div class="alert alert-info">
<strong>Recreate the plot below (or customize your own plot). </strong>
Note that the fill color is "#0080FF", and the shape is 21, and the theme is classic. The linewidth = 0.75, and linetype = 'dashed' .

```{r pq1c, fold.plot = F, echo = F, fig.height = 3, fig.width = 5, fig.align = 'center', eval = F}
pq <- 
  ggplot(data = visits, aes(x = Year, y = Annual_Visits/1000)) +
    geom_line(linewidth = 0.75, linetype = 'dashed') + 
    geom_point(color = "black", fill = "#0080FF", size = 2.5, shape = 21) +
    labs(x = "Year", y = "Annual visits in 1,000s") +
    scale_y_continuous(limits = c(2000, 4500),
                       breaks = seq(2000, 4500, by = 500)) + 
    scale_x_continuous(limits = c(1994, 2024),
                       breaks = c(seq(1994, 2024, by = 5))) + 
    theme_classic()
pq
```

<details><summary class='code2'>Answer</summary>
```{r pq1ac, eval = F, results = 'hide', eval = F}
ggplot(data = visits, aes(x = Year, y = Annual_Visits/1000)) +
  geom_line(linewidth = 0.75, linetype = 'dashed') + 
  geom_point(color = "black", fill = "#0080FF", size = 2.5, shape = 21) +
  labs(x = "Year", y = "Annual visits in 1,000s") +
  scale_y_continuous(limits = c(2000, 4500),
                     breaks = seq(2000, 4500, by = 500)) + 
  scale_x_continuous(limits = c(1994, 2024),
                     breaks = c(seq(1994, 2024, by = 5))) + 
  theme_classic()

```
</details>
</div>
<br>
</details>
<br>
