---
output: 
  html_document:
    css: custom_styles.css
---
#### Conditional Functions
<details open><summary class = 'drop'>Conditionals 101</summary>
Conditional functions `ifelse()`, `if(){ }else{ }`, and `case_when()` allow you to return results that depends on specified conditions. 

The main differences between the most common conditional functions:
<ul>
<li>`ifelse()`: Primarily for use with data frames. Takes 3 arguments: 1) the condition to test; 2) the value to return if condition is true; 3) the value to return of the condition is false. Function can only handle 2 possible outcomes, although nested `ifelse()` statements are possible (see example below). This function is vectorized, which means it's optimized for working on columns in data frames. Of the 3 conditionals, it tends to perform the fastest on large data sets.</li>
<li>`case_when()`: Primarily for use with data frames. Can take any number of condition statements and their value to return. Requires `dplyr` package to be loaded. Syntax is a bit tricky to figure out at first, but once you have it, it's about as easy as using `ifelse()`. This function is akin to SQL CASE WHEN. On large data sets, it consistently performs slower than `ifelse()`. </li>
<li>`if(){ }else{ }`: Can be used with data frames, but is more commonly used for operations outside of data frames. An example would be only running a chunk of code if a certain condition is met (e.g., if the data frame has > 0 rows, run next line of code.)</li>
</ul>

<details open><summary class = 'drop2'>`ifelse()`</summary>
The `ifelse()` function takes 3 arguments organized like: `ifelse(condition == TRUE, return this, return this instead)`. The first is the condition you're testing. The second argument is what to return if the condition is met. The third is what to return if the condition is not met. You can also nest `ifelse()` to include more than 2 conditions, but it can quickly get out of hand and hard to follow (see below).

Let's start by adding a column to the NETN tree data that uses the `TreeStatusCode` to create a new column called `status` that is either live or dead conditional on the abbreviated code in `TreeStatusCode` in `trees_final`.

```{r day2_cd, echo = F, include = F}
#------------------------------------
#      Day 2: Conditionals Code 
#------------------------------------
```

<p class = 'code'>Create status column conditioning on TreeStatusCode</p>
```{r statc, results = 'hide'}
# Check the levels of TreeStatusCode
sort(unique(trees_final$TreeStatusCode))
alive <- c("AB", "AL", "AS", "RS")
dead <- c("DB", "DM", "DS")

trees_final <- trees_final |> 
  mutate(status = ifelse(TreeStatusCode %in% alive, "live", "dead"))

# nested ifelse to make alive, dead, and recruit 
trees_final <- trees_final |> 
  mutate(status2 = ifelse(TreeStatusCode %in% dead, "dead",
                          ifelse(TreeStatusCode == "RS", "recruit", 
                                 "dead")))

```

Remember that we used `%in%` instead of `==` because alive has multiple status codes. We used `==` in `TreeStatusCode == "RS"` because there's only one status code considered a recruit. We could have also used `%in%` with the same results, but that would not be true for matching the alive status codes.
</details>
<br>

<details open><summary class = 'drop2'>`case_when()`</summary>
The `case_when()` function allows you to have multiple conditions, each with their own return. The syntax is a bit different than `ifelse()` to allow for the multiple conditions and returns. Using the same approach as above, we'll create a third status code with `case_when()`. We'll specify the output for when a tree status code is in the dead category, recruit, and live category. We'll then add a fourth output for status codes that don't match any of the previous conditions and set that as 'unknown'. Basically the `TRUE` just means, any records left are assigned 'unknown'. Note also the order of operations in `case_when()`. The alive group includes "RS", but is not assigned "live", because it was alreay matched. The `case_when()` function starts with the first statement at the top (i.e. dead trees). Any record that matches the first statement is then dropped from additional statements. The second statement considers all trees not matched as dead. The third statment considers all trees not matched as dead or recruit. Then the fourth statement considers any trees not matched as dead, recruit, or live. Rather than relying on this function behavior, it's better to not have overlapping categories (e.g. not include "RS" in alive). I include it here to demonstrate the point. 

<p class = 'code'>Create status column conditioning on TreeStatusCode</p>
```{r statc2}
# Check the levels of TreeStatusCode
alive <- c("AB", "AL", "AS", "RS")
dead <- c("DB", "DM", "DS")

trees_final <- trees_final |> 
  mutate(status3 = case_when(TreeStatusCode %in% dead ~ 'dead',
                             TreeStatusCode == 'RS' ~ 'recruit',
                             TreeStatusCode %in% alive ~ 'live', 
                             TRUE ~ 'unknown'))

table(trees_final$status2, trees_final$status3) # check that the output is the same

```
</details>
<br>
<details open><summary class = 'drop2'>`if(){ }else{ }`</summary>
This style of `if(){ }else{ }`, hereafter called if/else, conditionals is best used for operations outside of data frames, like turning code on or off based on specific conditions. I use if/else with ggplot (graphing R package we'll cover later) to turn certain features on or off based on a condition in the data or a condition I set. If/else statements are also helpful for bug handling in your code. For example, if you want the code to send a warning when your data frame is empty (no rows), you can have an if/else statement that prints to the console. You can string together multiple conditions to test by adding more `else{ }` statements.

<p class = 'code'>Print warning in console that indicates if invasive species are found in wetland data</p>
```{r ifelsebug1, fold.output = F}
inv <- ACAD_wetland |> filter(Invasive == TRUE)

if(nrow(inv) > 0){print("Invasive species were detected in the data.")
  } else {print("No invasive species were detected in the data.")}

```

<p class = 'code'>Force the else statement to print, by filtering out invasive species before testing. I added another potential else statment just to show that syntax.</p>
```{r ifelsebug2, fold.output = F}
native_only <- ACAD_wetland |> filter(Invasive == FALSE) 
inv2 <- native_only |> filter(Invasive == TRUE)

if(nrow(inv2) > 0){print("Invasive species were detected in the data.")
  } else if(nrow(inv2) == 0){print("No invasive species were detected in the data.")
    } else {"Invasive species detections unclear"}

```

</details>
</details>
<br>

<details open><summary class = "drop">Test your skills with conditionals</summary>

<div class="alert alert-info">
<strong>CHALLENGE: Using the ACAD_wetlands data, create a new column called Status that has "protected" for Protected = TRUE and "public" values for Protected = FALSE. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetcond, eval = F}
# read in wetland data if you don't already have it loaded.
ACAD_wetland <- read.csv("./data/ACAD_wetland_data_clean.csv")

# Base R using the with() function
ACAD_wetland$Status <- with(ACAD_wetland, ifelse(Protected == TRUE, "protected", "public"))
table(ACAD_wetland$Status, ACAD_wetland$Protected)
```

```{r wetcond2, eval = T}
# Tidyverse
ACAD_wetland <- ACAD_wetland |> mutate(Status = ifelse(Protected == TRUE, "protected", "public"))
table(ACAD_wetland$Status, ACAD_wetland$Protected) # check your work
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the ACAD_wetlands data, create a new column called abundance_cat that has levels High, Medium, Low, based on Ave_Cov, where "High" is >50%, "Medium" is 10-50%, and "Low" is < 10%. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetab, eval = F}
# Base R using the with() function and nested ifelse()
ACAD_wetland$abundance_cat <- with(ACAD_wetland, ifelse(Ave_Cov < 10, "Low",
                                                        ifelse(Ave_Cov >= 10 & Ave_Cov <= 50, "Medium", "High")))
```

```{r wetab2, eval = T}
# Tidyverse using case_when() and between
ACAD_wetland <- ACAD_wetland |> mutate(abundance_cat = case_when(Ave_Cov < 10 ~ "Low",
                                                                 between(Ave_Cov, 10, 50) ~ "Medium", 
                                                                 TRUE ~ "High"))
table(ACAD_wetland$abundance_cat)
```
Note the use of the `between()` function that saves typing. This function matches as >= and <=. 

</details>
</div>
<br>
</details>
<br>
