---
output: 
  html_document:
    css: custom_styles.css
---

#### Wrangling with dplyr 
<details open><summary class = 'drop'>Introduction to `dplyr` </summary>
The `dplyr` package is perhaps the single most useful package in R for working with your data. 
<img src = "./images/dplyr_filter.png" alt = 'dplyr filter' style = "float:right; width:375px;">
<span class = "capright"><a href="https://allisonhorst.com">Artwork by @allison_horst</a></span>

Commonly used `dplyr` functions and their use:
<ul>
<li><b>filter():</b> filters data for observations that meet specific criteria.</li>
<li><b>select():</b> subsets columns by either selecting or removing them.</li>
<li><b>arrange():</b> sorts data by specified column. </li>
<li><b>mutate():</b> adds new column(s) to a data frame. </li>
<li><b>slice():</b> slices the data based on specified number of rows</li>
<li><b>summarize(.by = groups):</b> summarizes by groups/factor levels and returns data for each group (e.g. mean cover grouped by plot).
<li><b>mutate(.by = groups):</b> summarizes by groups/factor levels and returns the original number of rows and columns.</li>
<li><b>rename():</b> renames columns</li>
</ul>

Now, using the `dplyr` package in the `tidyverse`, we're going to do the same operations we did yesterday with brackets. 

The steps we took with the Bass Harbor motile invertebrate data were: 
<ul>
<li>Convert Damage to a number.</li>
<li>Convert StartDate (character) to a Date type. </li>
<li>Rename the ScientificName column to Species.</li>
<li>Create a Site_Plot column that's a combination of SiteCode and PlotName.</li>
<li>Change No.Damage count that's 1960 to 196.</li>
<li>Drop records that are QAQC visits. </li>
</ul>
</details>
<br>

```{r day2_wra, echo = F, include = F}
#------------------------------------
#     Day 2: Data Wrangling Code 
#------------------------------------
```

<details open><summary class = 'drop'>Wrangle in dplyr</summary>
<ol>
<li><p class = 'code'>Read in example motile invertebrate data</p></li>
```{r readmotinv3b}
motinv <- read.csv("./data/BASHAR_motile_invert_counts.csv")
```

<li><p class = 'code'>Replace "PM" with NA (blank) in Damage columns.</p></li>
```{r pmfixd2, eval = F}
# Base R
motinv2 <- motinv
motinv2$Damage[motinv2$Damage == "PM"] <- NA
motinv2$Damage_num <- as.numeric(motinv2$Damage)
```
```{r pmfixbased2, fold.output = T}
# dplyr approach with mutate
motinv2 <- mutate(motinv, Damage_num = as.numeric(replace(Damage, Damage == "PM", NA)))
str(motinv2)
```

<li><p class = 'code'>Convert StartDate (character) to Date (date-time).</p></li>
```{r tr3dtd2, eval = F}
# Base R
motinv2$Date <- as.Date(motinv2$StartDate, format = "%m/%d/%Y")
```
```{r trdtd2dp, fold.output = T}
# dplyr approach with mutate
motinv2 <- mutate(motinv2, Date = as.Date(StartDate, format = "%m/%d/%Y"))
```

<li><p class = 'code'>Rename the ScientificName column to Species.</p></li>
```{r renamed2, eval = F}
# Base R code
names(motinv2)[names(motinv2) == "ScientificName"] <- "Species"
```
```{r renamedd2, fold.output = T}
# dplyr approach with rename
motinv2 <- rename(motinv2, "Species" = "ScientificName")
names(motinv2)
```

<li><p class ='code'>Create a Site_Plot column that's a combination of SiteCode and PlotName.</p></li>
```{r pasta2d2, eval = F}
# Base R
motinv2$Site_Plot <- paste(motinv2$SiteCode, motinv2$PlotName, sep = "-")
```
```{r pasta2bd2}
# dplyr approach with mutate
motinv2 <- mutate(motinv2, Site_Plot = paste(SiteCode, PlotName, sep = "-"))
```

<li><p class = 'code'>Drop records that are QAQC visits and drop original Damage column.</p></li>
```{r tr3d2, eval = F}
# Base R
motinv3 <- subset(motinv2, QAQC == FALSE, select = -Damage) # Note the importance of FALSE all caps
```
```{r tr3bd2, fold.output = T}
# dplyr
motinv3a <- filter(motinv2, QAQC == FALSE)
motinv3 <- select(motinv3a, -Damage)

head(motinv3)
```
</ol>
Note that subsetting data frames in R, which refers to reducing rows, columns, or both, is split between 2 functions in `dplyr`. The `filter()` function reduces rows. The `select()` function reduces columns. 
</details>
<br>

<details open><summary class = 'drop'>The magic pipe `|>`</summary>
The pipe (`|>` or `%>%`) makes `dplyr` and other tidyverse packages even more powerful. The pipe `|>` allows you to string together commands. So, taking all of the code above, we can do it all in the same function call. 

<p class='code'>Wrangle motile invertebrate data with pipes</p>
```{r pipe, fold.output = T}
motinv_final <- motinv |> 
  mutate(Damage_num = as.numeric(replace(Damage, Damage == "PM", NA)), # Fix Damage PM
         SitePlot = paste(SiteCode, PlotName, sep = "-"), # create new SitePlot column
         Date = as.Date(StartDate, format = "%m/%d/%Y")) |> # create new Date column
  rename("Species" = "ScientificName") |> # change column name
  filter(QAQC == FALSE) |> # drop QAQC visits
  select(-Damage) |> # drop original Damage column
  arrange(SitePlot, Year, Species) # optional sorting the data

head(motinv_final)  
```

Hopefully you agree that pipes are amazing! They allow for more efficient coding and in relatively easy to follow the steps, and make the dplyr functions, like `mutate()` so much more useful. Outside of pipes for example, `mutate()` doesn't feel more useful than base R for creating a new column. From now on, I will use pipes regularly in the code. 

If you've ever seen the `%>%`, that also functions as a pipe with code. The `%>%` pipe was the original pipe that was introduced by the tidyverse in the `magrittr` package. The `magrittr` pipe was so popular, that starting in R 4.0, a base R pipe was introduced (`|>`). It's supposed to be better optimized for order of operations and reduces a package you need to install. So, in general, use the base R pipe `|>`. It's also why I had you set the default pipe in Global Options to the `|>`. A useful keyboard shortcut for the pipe is Ctrl + Shift + M. You should see the `|>` pipe in your script when you type that shortcut. If you get the `%>%` pipe instead, you need to change that default setting in Global Options (see Day 1 > R and RStudio > RStudio Global Options > Step 3. Change default pipe.)
<br>
<br>
<p class = 'note'><b>Coding Tip:</b> While the number of steps you can pipe together is virtually endless, piping many tasks, especially complex ones, can make code hard to read and troubleshoot. It's best to limit number of pipes to 3-4, and/or to do complex tasks that might fail or require checking on their own.</p>
</details>
<br>

<details open><summary class = 'drop'>Test your skills with dplyr!</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using `motinv`, how many species are found in PlotName A1 in 2024?</strong>
<details><summary class='drop2'>Answer</summary>

```{r chal4dp}
# with brackets
A1_2024 <- motinv |> filter(PlotName == "A1" & Year == 2024)
nrow(A1_2024) # 3
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What years have green crabs (Latin: Carcinus maenas) been detected? </strong>
<details><summary class='drop2'>Answer</summary>

```{r chal5bdp, fold.output = F}
gcrab <- motinv |> filter(ScientificName == "Carcinus maenas") |> 
  select(Year) |> unique()

gcrab
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find the highest value recorded in No.Damage column.</strong>
<details><summary class='drop2'>Answer</summary>
```{r maxnddp, fold.output = F}
max_nd <- max(motinv$No.Damage, na.rm = TRUE)

motinv |> filter(No.Damage == max_nd)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Fix the No.Damage typo by replacing 1960 with 196.</strong>

<details><summary class = 'drop2'>Answer</summary>
Let's say that you looked at the datasheet, and the actual count for No.Damage was 196 instead of 1960. You can change that value in the original CSV by hand. But even better is to document that change in code. There are multiple ways to do this. Two examples are below.

But first, it's good to create a new data frame when modifying the original data frame, so you can refer back to the original if needed. I also use a really specific filter to make sure I'm not accidentally changing other data.

<p class = 'code'>Replace 1960 with 196</p>
```{r trfxd2t2, fold.output = F}
# Reminder of the base R approach

# create copy of motinv data
motinv_fix <- motinv

# find the problematic value, and change it to 196
motinv_fix$No.Damage[motinv_fix$Year == 2019 & 
                     motinv_fix$PlotName == "R4" & 
                     motinv_fix$No.Damage == 1960] <- 196

# dplyr approach
motinv_fix <- motinv |> mutate(No.Damage = replace(No.Damage, No.Damage == 1960, 196))
range(motinv_fix$No.Damage)
```
</details>
</div>
</details>
<br>