---
output: 
  html_document:
    css: custom_styles.css
---

#### Summarizing with dplyr
<details open><summary class = 'drop'>Using `summarize()`</summary>
Yesterday, we used functions like `mean()`, `min()`, and `max()` to summarize entire datasets. Now we're going to use those same functions to summarize data by grouping variables, such as park, year, plot, etc. The process is similar to using Totals in Access or subtotals in Excel, although it is more flexible and efficient in R.

Difference between `summarize()` and `mutate()`:
<ul>
<li>`mutate()` returns the same number of <b>rows</b> as the original data frame. This function also returns all of the rows that were in the original data frame.</li>
<li>`summarize()` returns the same number of rows as there are <b>grouping levels</b> in the original data frame. This function only returns the rows that were part of the `.by = c()` and that were created in the `summarize()` function. 
</ul>

Common functions used to summarize:
<ul>
<li>`mean()`: calculate the group means</li>
<li>`min()`: calculate the group minimums</li>
<li>`max()`: calculate the group maximums</li>
<li>`sum()`: calculate the group sums</li>
<li>`sd()`: calculate the group standard deviations</li>
<li>`n()`: tally the number of rows within each group</li>
</ul>

To demonstrate summarize and mutate in dplyr, we're going to use the point intercept data collected along 3 transects in the Bass Harbor site. The data have already been summarized by transect (T1, T2, T3). We now want to calculate site-level median elevation and percent frequency for each cover type.

<p class = 'code'>Read in the point intercept data </p>

```{r day2_gs, echo = F, include = F}
#------------------------------------
#      Day 2: Summarizing Code 
#------------------------------------
```

```{r pid2s}
pi_dat <- read.csv("./data/BASHAR_Point_Intercept_data.csv")

head(pi_dat)
```

<p class = 'code'>Using `mutate()`, calculate the average percent frequency and median elevation by CoverType and Year</p>
```{r ttrsummut, message = F, warning = F, eval = F}
pi_dat_mut <- pi_dat |> mutate(med_elev_sl = median(med_elev), 
                               avg_pct_freq = mean(pct_freq), 
                               .by = c(SiteCode, Year, CoverType, CoverCode))
nrow(pi_dat) #314
nrow(pi_dat_mut) #314
head(pi_dat_mut)
```
Note how `pi_dat_mut` has the same number of rows and all the original columns plus the two we calculated (site-level median, and average % frequency). More often we're interested in reducing the data to one row per grouping level. That's what `summarize()` is for.

<p class = 'code'>Using `summarize()`, calculate the average percent frequency, median elevation, min/max of frequency and elevation by CoverType and Year.</p>
```{r ttrsummut2, message = F, warning = F, eval = F}
pi_dat_sum <- pi_dat |> summarize(elev_sl_med = median(med_elev), 
                                  elev_sl_min = min(med_elev),
                                  elev_sl_max = max(med_elev),
                                  avg_pct_freq = mean(pct_freq), 
                                  .by = c(SiteCode, Year, CoverType, CoverCode))
nrow(pi_dat) #314
nrow(pi_dat_sum) #124
head(pi_dat_sum)
```
Note how `pi_dat_sum` has the 1/3 of the rows and only the grouping columns and the two we calculated (site-level median, and average % frequency).

The `mutate(.by = c())` approach is helpful if you're trying to standardize values within your group. But in most cases, the `summarize(.by = c())` approach, which collapses to the group level, is what you're looking for. Note that in older versions of dplyr, the syntax was `group_by() |> summarize()` with no `.by = c()`.

<p class = 'code'>Summarize the average and standard error total counts for the motile invertebrate data by CommunityType.</p>
We will first fix the 1960 and PM errors from before, drop QAQC visits, then combine the Damage and No.Damage columns to make a total_count column.
```{r trdbhsum, eval = F}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))

head(motinv_sum)
```
</details>
<br>

<details open><summary class = "drop">Test your summarizing skills</summary>

<div class="alert alert-info">
<strong>CHALLENGE: Using the point intercept data (`pi_dat`), calculate the average percent frequency of each non-vegetated substrate by year. Note that non-vegetated substrates are CoverCode = c('BOLT', 'ROCK', 'WATER'). </strong>
<details><summary class='code2'>Answer</summary>
```{r sumnv, eval = F}
pi_nonveg <- pi_dat |> filter(CoverCode %in% c("BOLT", "ROCK", "WATER")) |> # filter nonveg grps
  summarize(avg_freq = mean(pct_freq), # calc avg.
            .by = c(SiteCode, Year, CoverCode, CoverType)) # grouping variables 

head(pi_nonveg) # check output
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the point intercept data (`pi_dat`), calculate the average percent frequency of each non-vegetated vs vegetated cover types by year. Note that non-vegetated substrates are CoverCode = c('BOLT', 'ROCK', 'WATER'). </strong>
<details><summary class='code2'>Answer</summary>
```{r sumnv2, eval = F}
pi_subtype <- pi_dat |>
  mutate(sub_type = ifelse(CoverCode %in% c("BOLT", "ROCK", "WATER"), "nonveg", "veg")) |> # filter nonveg grps
  summarize(avg_freq = mean(pct_freq), # calc avg.
            .by = c(SiteCode, Year, sub_type)) |> # grouping variables 
  arrange(SiteCode, Year, sub_type) # sort variables

head(pi_subtype) # check output
```
</details>
</div>
</details>
<br>
