---
output: 
  html_document:
    css: custom_styles.css
---
#### Conditional Functions
<details open><summary class = 'drop'>Conditionals 101</summary>
Conditional functions `ifelse()`, `if(){ }else{ }`, and `case_when()` allow you to return results that depends on specified conditions. 

The main differences between the most common conditional functions:
<ul>
<li>`ifelse()`: Primarily for use with data frames. Takes 3 arguments: 1) the condition to test; 2) the value to return if condition is true; 3) the value to return of the condition is false. Function can only handle 2 possible outcomes, although nested `ifelse()` statements are possible (see example below). This function is vectorized, which means it's optimized for working on columns in data frames. Of the 3 conditionals, it tends to perform the fastest on large data sets.</li>
<li>`case_when()`: Primarily for use with data frames. Can take any number of condition statements and their value to return. Requires `dplyr` package to be loaded. Syntax is a bit tricky to figure out at first, but once you have it, it's about as easy as using `ifelse()`. This function is akin to SQL CASE WHEN. On large data sets, it consistently performs slower than `ifelse()`. </li>
<li>`if(){ }else{ }`: Can be used with data frames, but is more commonly used for operations outside of data frames. An example would be only running a chunk of code if a certain condition is met (e.g., if the data frame has > 0 rows, run next line of code.)</li>
</ul>

<details open><summary class = 'drop2'>`ifelse()`</summary>
The `ifelse()` function takes 3 arguments organized like: `ifelse(condition == TRUE, return this, return this instead)`. The first is the condition you're testing. The second argument is what to return if the condition is met. The third is what to return if the condition is not met. You can also nest `ifelse()` to include more than 2 conditions, but it can quickly get out of hand and hard to follow (see below).

Let's start by adding a column to the motile invertebrate that uses the `SpeciesCode` to create a new column called `native` that is either TRUE for native species, or FALSE for non-native species. We'll add a second column named `native_grp` that is "native", "exotic", and "invasive". The invasive group includes Asian shore crabs and green crabs.

```{r day2_cd, echo = F, include = F}
#------------------------------------
#      Day 2: Conditionals Code 
#------------------------------------
```

<p class = 'code'>Create nativity column conditioning on SpeciesCode</p>
```{r statc,}
# green crab, Asian shore crab, and common periwinkle species codes
exo_spp <- c("CARMAE", "HEMISAN", "LITLIT") 

# smooth periwinkle, rough periwinkle, dogwhelk, and limpet species codes
nat_spp <- c("LITOBT", "LITSAX", "NUCLAP", "TECTES")

# Make a table of species codes in BASHAR
table(motinv$SpeciesCode)

# Add native column with ifelse
motinv <- motinv |> mutate(native = ifelse(SpeciesCode %in% nat_spp, TRUE, FALSE))

# Add native_status column with nested ifelse
motinv <- motinv |> mutate(native_status = ifelse(SpeciesCode %in% nat_spp, "native",
                                                  ifelse(SpeciesCode %in% c("CARMAE", "HEMISAN"), "invasive", 
                                                         "exotic")))

table(motinv$SpeciesCode, motinv$native)
table(motinv$SpeciesCode, motinv$native_status)
```

Remember that we used `%in%` instead of `==` because exotic has multiple species codes. Only use `==` when there's only one condition to match against. The `%in%` approach can with with any combination of matching conditions, so I almost always use `%in%` instead of `==`.
</details>
<br>

<details open><summary class = 'drop2'>`case_when()`</summary>
The `case_when()` function allows you to have multiple conditions, each with their own return. The syntax is a bit different than `ifelse()` to allow for the multiple conditions and returns. Using the same approach as above, we'll recreate the native_status column with `case_when()`. We'll then add a fourth output for species codes that don't match any of the previous conditions and set that as 'unknown'. Basically the `TRUE` just means, any records left are assigned 'unknown'. 

Note the order of operations in `case_when()`. The first step assigns native species a 'native' status. Then, only non-native species are left to condition on. The next step assigns CARMAE and HEMISAN as 'invasive'. The third step conditions on species on the exo_spp group, but only those that weren't already handled in previous steps. Then the fourth statement considers any species not matched as native, invasive, exotic. Rather than relying on this function behavior, it's better to not have overlapping categories (e.g. not include CARMAE and HEMISAN in the exo_spp). I include it here to demonstrate the point. 

<p class = 'code'>Create status column conditioning on SpeciesCode</p>
```{r statc2}
# green crab, Asian shore crab, and common periwinkle species codes
exo_spp <- c("CARMAE", "HEMISAN", "LITLIT") 

# smooth periwinkle, rough periwinkle, dogwhelk, and limpet species codes
nat_spp <- c("LITOBT", "LITSAX", "NUCLAP", "TECTES")

motinv <- motinv |> 
  mutate(native_status = case_when(SpeciesCode %in% nat_spp ~ 'native',
                                   SpeciesCode %in% c("CARMAE", "HEMISAN") ~ 'invasive',
                                   SpeciesCode %in% exo_spp ~ 'exotic', 
                                   TRUE ~ 'unknown'))

table(motinv$SpeciesCode, motinv$native_status) # check that the output worked

```
</details>
<br>
<details open><summary class = 'drop2'>`if(){ }else{ }`</summary>
This style of `if(){ }else{ }`, hereafter called if/else, conditionals is best used for operations outside of data frames, like turning code on or off based on specific conditions. I use if/else with ggplot (graphing R package we'll cover later) to turn certain features on or off based on a condition in the data or a condition I set. If/else statements are also helpful for bug handling in your code. For example, if you want the code to send a warning when your data frame is empty (no rows), you can have an if/else statement that prints to the console. You can string together multiple conditions to test by adding more `else{ }` statements.

<p class = 'code'>Print warning in console that indicates if invasive species are found in the motile invertebrate data.</p>
```{r ifelsebug1, fold.output = F}
inv <- motinv |> filter(native_status == "invasive")
spp_det <- unique(inv$CommonName)

if(nrow(inv) > 0){print(paste0("The following invasive species were detected in the data: ", 
                               paste0(spp_det, collapse = ", ")))
  } else {print("No invasive species were detected in the data.")}

```

<p class = 'code'>Force the else statement to print, by filtering out invasive species before testing. I added another potential else statement just to show that syntax.</p>
```{r ifelsebug2, fold.output = F}
inv <- motinv |> filter(SpeciesCode %in% nat_spp) |> 
  filter(native_status == "invasive")
spp_det <- unique(inv$CommonName)

if(nrow(inv) > 0){print(paste0("The following invasive species were detected in the data: ", 
                               paste0(spp_det, collapse = ", ")))
  } else {print("No invasive species were detected in the data.")}
```

</details>
</details>
<br>

<details open><summary class = "drop">Test your skills with conditionals</summary>

<div class="alert alert-info">
<strong>CHALLENGE: Using the motinv data, create a new column called trophic that indicates whether the species is an herbivore or predator. <br>
Hint: predator site codes are c("CARMAE", "NUCLAP"), and herbivore site codes are c("LITLIT", "LITOBT", "LITSAX", "TECTES"). </strong>
<details><summary class='code2'>Answer</summary>
```{r wetcond}
pred <- c("CARMAE", "NUCLAP")

# base R
motinv$trophic <- ifelse(motinv$SpeciesCode %in% pred, "predator", "herbivore")
table(motinv$trophic, motinv$SpeciesCode)

# tidyverse
motinv <- motinv |> mutate(trophic = ifelse(SpeciesCode %in% pred, "predator", "herbivore"))
table(motinv$trophic, motinv$SpeciesCode)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the motile invertebrate data, create a new column called count_level that has levels High, Medium, Low, based on No.Damage, where "High" is > 35, "Medium" is 10 - 35, and "Low" is < 10. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetab}
# Base R using a nested ifelse()
motinv$count_level <- 
  ifelse(motinv$No.Damage > 35, "High", 
         ifelse(motinv$No.Damage >= 10 & motinv$No.Damage <= 35, "Medium", "Low"))

table(motinv$count_level) # check that it worked
```

```{r wetab2}
# Tidyverse using case_when() and between
motinv <- motinv |> mutate(count_level = case_when(No.Damage > 35 ~ "High",
                                                   between(No.Damage, 10, 35) ~ "Medium",
                                                   No.Damage < 10 ~ "Low"))

table(motinv$count_level) # check that it worked
```
Note the use of the `between()` function that saves typing. This function matches as >= and <=. 

</details>
</div>
<br>
</details>
<br>
