---
output: 
  html_document:
    css: custom_styles.css
---

#### Summarizing with dplyr
<details open><summary class = 'drop'>Using `summarize()`</summary>
Yesterday, we used functions like `mean()`, `min()`, and `max()` to summarize entire datasets. Now we're going to use those same functions to summarize data by grouping variables, such as park, year, plot, etc. The process is similar to using Totals in Access or subtotals in Excel, although it is more flexible and efficient in R.

Difference between `summarize()` and `mutate()`:
<ul>
<li>`mutate()` returns the same number of <b>rows</b> as the original data frame. This function also returns all of the rows that were in the original data frame.</li>
<li>`summarize()` returns the same number of rows as there are <b>grouping levels</b> in the original data frame. This function only returns the rows that were part of the `.by = c()` and that were created in the `summarize()` function. 
</ul>

Common functions used to summarize:
<ul>
<li>`mean()`: calculate the group means</li>
<li>`min()`: calculate the group minimums</li>
<li>`max()`: calculate the group maximums</li>
<li>`sum()`: calculate the group sums</li>
<li>`sd()`: calculate the group standard deviations</li>
<li>`n()`: tally the number of rows within each group</li>
</ul>

To demonstrate summarize and mutate in dplyr, we're going to use the point intercept data collected along 3 transects in the Bass Harbor site. The data have already been summarized by transect (T1, T2, T3). We now want to calculate site-level median elevation and percent frequency for each cover type.

<p class = 'code'>Read in the point intercept data </p>

```{r day2_gs, echo = F, include = F}
#------------------------------------
#      Day 2: Summarizing Code 
#------------------------------------
```

```{r pid2s, eval = F}
pi_dat <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_Point_Intercept_data.csv"
)
head(pi_dat)
```

```{r pid2sl, echo = F}
pi_dat <- read.csv("./data/BASHAR_Point_Intercept_data.csv")
```

<p class = 'code'>Calculate the average percent frequency and median elevation by CoverType using mutate</p>
```{r ttrsummut, message = F, warning = F, eval = F}
pi_dat_mut <- pi_dat |> mutate(med_elev_sl = median(med_elev), 
                               avg_pct_freq = mean(pct_freq), 
                               .by = c(Year, CoverType, CoverCode))
nrow(pi_dat) #317
nrow(pi_dat_mut) #317
head(pi_dat_mut)
```
Note how `pi_dat_mut` has the same number of rows and all the original columns plus the two we calculated (site-level median, and average % frequency).

<p class = 'code'>Calculate the average percent frequency and median elevation by CoverType using summarize</p>
```{r ttrsummut2, message = F, warning = F, eval = F}
pi_dat_sum <- pi_dat |> summarize(med_elev_sl = median(med_elev), 
                                  avg_pct_freq = mean(pct_freq), 
                                  .by = c(Year, CoverType, CoverCode))
nrow(pi_dat) #317
nrow(pi_dat_sum) #127
head(pi_dat_sum)
```
Note how `pi_dat_sum` has the 1/3 of the rows and only the grouping columns and the two we calculated (site-level median, and average % frequency).

The `mutate(.by = c())` approach is helpful if you're trying to standardize values within your group. But in most cases, the `summarize(.by = c())` approach, which collapses to the group level, is what you're looking for. Note that in older versions of dplyr, the syntax was `group_by() |> summarize()` with no `.by = c()`.

<p class = 'code'>Summarize the average and standard error total counts for the motile invertebrate data by CommunityType.</p>
We will first fix the 1960 and PM errors from before, drop QAQC visits, then combine the Damage and No.Damage columns to make a total_count column.
```{r trdbhsum, eval = F}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))

head(motinv_sum)
```
</details>
<br>

<details open><summary class = "drop">Test your summarizing skills</summary>

#+++++ ENDED HERE ++++++

<div class="alert alert-info">
<strong>CHALLENGE: Using the ACAD_wetland data, calculate the average percent cover of native vs. invasive species per plot. Note that Invasive = TRUE is invasive and FALSE is native. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetinvcov, eval = F}
# Using group_by()
ACAD_inv <- ACAD_wetland |> group_by(Site_Name, Year, Invasive) |> 
  summarize(Pct_Cov = sum(Ave_Cov), 
            .groups = 'drop') |>  # optional line to keep console from being chatty
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_inv)

# Using summarize(.by)
ACAD_inv2 <- ACAD_wetland |> 
  summarize(Pct_Cov = sum(Ave_Cov), .by = c(Site_Name, Year, Invasive)) |> 
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_inv2) # should be the same as ACAD_inv
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the ACAD_wetland data, count the number of native vs. invasive species per plot. Note that Invasive = TRUE is invasive and FALSE is native. </strong>
<details><summary class='code2'>Answer</summary>
```{r wetinvr, eval = F}
# Using group_by()
ACAD_spp <- ACAD_wetland |> group_by(Site_Name, Year, Invasive) |> 
  summarize(num_spp = n(), 
            .groups = 'drop') |>  # optional line to keep console from being chatty
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_spp)

# Using summarize(.by)
ACAD_spp2 <- ACAD_wetland |> 
  summarize(num_spp = n(), .by = c(Site_Name, Year, Invasive)) |> 
  arrange(Site_Name) # sort by Site_Name for easier comparison

head(ACAD_spp2) # should be the same as ACAD_inv
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using ACAD_wetland data, calculate relative % cover of species within each site. This one is challenging!</strong>
<details><summary class='code2'>Answer</summary>

<p class = 'code'>First sum site-level cover using mutate to return a value for every original row.</p>
```{r wetrelcov, eval = F}
ACAD_wetland <- ACAD_wetland |> group_by(Site_Name, Year) |> 
  mutate(Site_Cover = sum(Ave_Cov)) |> 
  ungroup() # good practice to ungroup after group.

table(ACAD_wetland$Site_Name, ACAD_wetland$Site_Cover) # check that each site has a unique value.
head(ACAD_wetland)
```

<p class = 'code'>Next calculate relative cover grouped on Site_Name, Year, Latin_Name, and Common </p> 
```{r wetrelcov2, eval = F}
# Create new dataset because collapsing rows on grouping variables
# Using group_by() and summarize()
ACAD_wetland_relcov <- ACAD_wetland |> group_by(Site_Name, Year, Latin_Name, Common) |> 
  summarize(rel_cov = (Ave_Cov/Site_Cover)*100,
            .groups = 'drop') |> 
  ungroup()

```

<p class = 'code'>Check that relative cover sums to 100% within each site</p>
```{r wetrelcovcheck, eval = F}
# Using summarize(.by = ) 
ACAD_wetland_relcov2 <- ACAD_wetland |> #group_by(Site_Name, Year, Latin_Name, Common) |> 
  summarize(rel_cov = Ave_Cov/Site_Cover, 
            .by = c("Site_Name", "Year", "Latin_Name", "Common"))

# Check that your relative cover sums to 100 for each site
relcov_check <- ACAD_wetland_relcov2 |> group_by(Site_Name, Year) |> 
  summarize(tot_relcov = sum(rel_cov)*100, .groups = 'drop')
table(relcov_check$tot_relcov) # they should all be 100

```
</details>
</div>

</details>
<br>
