---
output: 
  html_document:
    css: custom_styles.css
---

#### Pivoting Tables
<details open><summary class='drop'>Reshaping 101</summary>
Reshaping data from long to wide and wide to long is a common task with our data.  Datasets are usually described as long, or wide. The long form, which is the structure database tables often take, consists of each row being an observation, and each column being a variable (i.e. tidy format). However, in summary tables, we often want to reshape the data to be wide for better digestion.  

We’ll work with the point intercept data again to demonstrate pivoting, and will use the data frame we created by summarizing median elevation and average percent frequency of each cover code by year. If you don't have that data frame yet, run the code below.  

```{r day3_pv, echo = F, include = F}
#------------------------------------
#         Day 3: Pivot Code 
#------------------------------------
```

```{r impbat, eval = T}
# load the package
library(dplyr)
library(tidyr) # for pivot functions

# import the raw point intercept data
pi_dat <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_Point_Intercept_data.csv"
)

# summarize data by site, year, and cover type
pi_dat_sum <- pi_dat |> summarize(med_elev_sl = median(med_elev, na.rm = T), 
                                  avg_pct_freq = mean(pct_freq, na.rm = T), 
                                  .by = c(SiteCode, Year, CoverType, CoverCode))

```

</details>
<br>

<details open><summary class='drop'>Pivot from long to wide</summary>
With `pi_dat_sum` we're going to pivot the data wide to make each `CoverCode` a separate column and the values in each cell be the avg_pct_freq. The code below is pretty straightforward with `names_from` being the column you want to turn into column names, and the `values_from` being the value you want in the cells. 

<p class = 'code'>Pivot point intercept data to wide</p>
When pivoting long to wide, you're reducing the number of rows to have one observation for each of level of the variable you're pivoting on. If you have other variables in your data frame, like CoverCode and med_elev_sl in this data frame, you have to drop those columns for the pivot to result in one observation per level of the pivoted variable. If that doesn't make sense, try running the `pivot_wider()` without dropping CoverType or med_elev_sl, and you'll see what I mean. I also added the `arrange()` by CoverCode, so the columns were sorted alphabetically in the pivot.
```{r piwide}
pi_wide <- pi_dat_sum |> 
  arrange(CoverCode, Year) |> # sort by CoverCode and year
  select(-CoverType, -med_elev_sl) |> # Drop extra column
  pivot_wider(names_from = CoverCode, # column that will produce column names
              values_from = avg_pct_freq) # column to make the values
head(pi_wide)
```

That was pretty simple. But there are a lot of blanks where a CoverCode wasn't detected in a give year and site. We can use the `values_fill` argument to save us time filling blanks as 0s.

<p class = 'code'>Pivot point intercept data to wide filling blanks as 0</p>
```{r piwide2}
pi_wide <- pi_dat_sum |> 
  arrange(CoverCode, Year) |> 
  select(-CoverType, -med_elev_sl) |> 
  pivot_wider(names_from = CoverCode, 
              values_from = avg_pct_freq, 
              values_fill = 0) # new line

head(pi_wide)
```
Now we see that every cell has a value. Another useful argument in `pivot_wider()` is `names_prefix`. That allows you to add a string before the column names that are generated in the pivot. This is helpful if you're pivoting on a number column, like year or plot number. R doesn't like column names that start with a number. The names_prefix is a quick way to fix that. To demonstrate, I'll pivot on year instead of CoverCode.

<p class = 'code'>Pivot point intercept data wide using Year instead of CoverCode and add prefix.</p>
```{r batwide3}
pi_wide_yr <- pi_dat_sum |> 
  arrange(Year) |>
  select(-med_elev_sl) |> 
  pivot_wider(names_from = Year, # pivot on year instead of CoverCode 
              values_from = avg_pct_freq, 
              values_fill = 0,
              names_prefix = "yr_") # new line

head(pi_wide_yr)
```

<div class="alert alert-info">
<strong>CHALLENGE: Use the `motinv_sum` data frame from the "Summarizing with dplyr" tab to pivot on SpeciesCode and mean_count, and fill the NAs with 0s. If you don't have the motinv_sum data frame handy, run the code below to create it. <br>
Hint: Drop the ScientificName and CommonName columns before you pivot.
</strong>

```{r}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))
```

<details><summary class='code2'>Answer</summary>
```{r pivwide2b, fold.output = F}
motinv_wide <- motinv_sum |> 
  arrange(SpeciesCode) |> # sorting so columns are alphabetical 
  select(-ScientificName, -CommonName) |> 
  pivot_wider(names_from = SpeciesCode,
              values_from = mean_count, 
              values_fill = 0)

head(motinv_wide)

```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Use the `motinv_sum` data frame from the "Summarizing with dplyr" tab to pivot on Year and mean_count, fill the NAs with 0s, and add "yr_" to the column names to prevent column names starting with numbers. If you don't have the motinv_sum data frame handy, run the code below to create it. <br>
Hint: Drop the se_counts column before you pivot.
</strong>

```{r moinvredo}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))
```

<details><summary class='code2'>Answer</summary>
```{r pivwide2c, fold.output = F}
motinv_wide_yr <- motinv_sum |> 
  arrange(Year) |> # sorting so columns are alphabetical 
  select(-se_counts) |> 
  pivot_wider(names_from = Year,
              values_from = mean_count, 
              values_fill = 0, 
              names_prefix = "yr_")

head(motinv_wide_yr)
```
</details>
</div>

</details><br>

<details open><summary class = 'drop'>Pivot wide to long</summary>
We can reshape the capture data back to long, which will give us a similar data as before with 0s are added into the data. For the `pivot_long()` function, you have to tell it which columns to pivot on. If you don't specify, it will make the entire dataset into 2 long columns, which you typically don't want. Here I tell R not to pivot on SiteCode and Year columns, because I know they're in the data frame and unlikely to change. If I instead specified the species codes to pivot on, if a new species were found in the next year of sampling, I'd have to update this code to include that new species.

```{r batlong}
pi_long <- pi_wide |> pivot_longer(cols = -c(SiteCode, Year), 
                                   names_to = "SpeciesCode", 
                                   values_to = "Avg_Pct_Freq")
head(pi_long)
```

Note that for `pivot_longer()` the `names_prefix = ""` argument actually removes the string you specify from the columns you're pivoting on, rather than adding the string to the column name in `pivot_wider()`. In other words, it does the opposite. 

<div class="alert alert-info">
<strong>CHALLENGE: Pivot the motinv_wide_yr data frame on the years columns, and remove the "yr_" from the year names using `names_prefix = 'yr_'`. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2a, fold.output = F}
motinv_long_yr <- pivot_longer(motinv_wide_yr, 
                               cols = -c(SiteCode, CommunityType, ScientificName, 
                                         CommonName, SpeciesCode),
                               names_to = "Year", 
                               values_to = "mean_counts", 
                               names_prefix = "yr_") # drops this string from values
```
</details>
</div>
</details>
<br>

#### Joining Tables
<details open><summary class='drop'>Joining tables 101</summary>
We often need to combine data from separate tables in our work (e.g., relational database tables). In R we do this using either the `merge()` function in base R or `join_()` functions in dplyr. Because I find dplyr join functions to be more intuitive and to perform faster than base R's merge, I'm going to show how to use dplyr. If you understand the basic concepts if the join functions, you can figure out how to merge in base R.

Joining tables requires that the two datasets to join have at least one column in common, which is referred to as the key. The key is used to match records. The join type will determine whether all rows from both datasets are returned, or if only a portion are returned based on values in either or both of the two datasets. We think of joining as consisting of a left and right dataset. The left is specified first in the function argument, and the right is specified second. It generally doesn't matter which you make left or right, just that you know which is left or right. In general, I put site-level datasets on the left, and sample event datasets on the right.  
</details><br>

<details open><summary class='drop'>Types of joins</summary>
<ul>
<li><b>Full Join</b>: keeps all observations that appear in both the left and right dataset. Any key value in the left dataset that is not found in the right dataset will return NAs for columns coming from the right data frame, and vice versa. The full join treats both the left and right datasets equally.</li>

<img src = "./images/full_join_R4DS.png" alt = 'full join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>
<br>
<li><b>Inner Join</b>: keeps only observations that are matched in both the left and right dataset. Any key value in the left dataset that is not found in the right dataset will be dropped, and vice versa. The inner join treats the left and right datasets equally. </li>

<img src = "./images/inner_join_R4DS.png" alt = 'inner join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>
<br>
<li><b>Left Join</b>: keeps all observations that appear in the left dataset (first specified) and only those matched in the right dataset. Any key value in the left dataset that is not found in the right dataset will return NAs for columns coming from the right dataset. Any key value in the right dataset not found in the left data frame will be dropped. </li>

<img src = "./images/left_join_R4DS.png" alt = 'left join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>
<br>
<li><b>Right Join</b>: keeps all observations that appear in the right dataset (second specified) and only those matched in the left dataset. Any key value in the right dataset that is not found in the left dataset will return NAs for columns coming from the left dataset. Any key value in the left dataset not found in the right dataset will be dropped. </li>

<img src = "./images/right_join_R4DS.png" alt = 'right join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>

<li><b>Anti Join</b>: Returns records in the left dataset that are not found in the right dataset. This is one direction, so to find all records not in common, you have to do an anti join with both combinations of data being on left or right side. I use anti joins when I'm trying to check that two datasets have the same sites or years represented.</li> 

<img src = "./images/anti_join_R4DS.png" alt = 'anti join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>
</ul>
</details><br>

<details open><summary class = 'drop'>Joins in practice</summary>
To demonstrate the different joins, we're going to use some fake bat capture data. One table has species captured by year and site. The other has additional site information, like X/Y coordinates, full site names, etc. 

```{r day3_jn, echo = F, include = F}
#------------------------------------
#        Day 3: Join Code 
#------------------------------------
```

<p class='code'>Read in motile site data </p>
```{r batsite, fold.output = F, eval = T}
# Import site data
bat_sites <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/bat_site_info.csv")

# Import bat capture data
bat_cap <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/bat_captures.csv")

# View sites listed in each
sort(unique(bat_sites$Site)) # Sites 1, 2, 3, 4, 5
sort(unique(bat_cap$Site)) # Sites 1, 2, 3, 5, 6
```

The key in the two bat datasets is the "Site" column. In the `bat_sites` data frame, there are 5 unique sites, numbered 1:5. In the `bat_cap` data there are 5 unique sites, numbered 1, 2, 3, 5, 6. Therefore site_004 is only found in `bat_sites` and site_006 is only found in `bat_cap`.  

<p class='code'>Full join</p>
```{r fullj, fold.output = F}
bat_full <- full_join(bat_sites, bat_cap, by = "Site")
table(bat_full$Site)
```

```{r fulljk, echo = F}
knitr::kable(bat_full, align = 'c') |> 
  kableExtra::scroll_box(height = "300px") |> 
  kableExtra::kable_styling(full_width = F, html_font = 'Arial', font_size = 10) |> 
  kableExtra::column_spec(1:10, background = 'white', include_thead = T)
```

Note how site_004, which was not in the `bat_cap` capture data, but was in the `bat_site` data is included with NAs for the columns that came from the `bat_cap` data. Additionally, site_006, which was only in the `bat_cap` capture data but not in the `bat_site` data has NAs for the columns that came from the `bat_site` data. 

<p class='code'>Inner join</p>
```{r innerj, fold.output = F}
bat_inner <- inner_join(bat_sites, bat_cap, by = "Site")
table(bat_inner$Site)
```

```{r innerjk, echo = F}
knitr::kable(bat_inner) |> 
  kableExtra::scroll_box(height = "300px") |> 
  kableExtra::kable_styling(full_width = F, html_font = 'Arial', font_size = 10) |> 
  kableExtra::column_spec(1:10, background = 'white', include_thead = T)
```

The inner join only returns records from both datasets that have site in common. Therefore, site_004 in the `bat_site` data and site_006 in the `bat_cap` capture data were dropped.

<p class='code'>Left join</p>
```{r leftj, fold.output = F}
bat_left <- left_join(bat_sites, bat_cap, by = "Site")
table(bat_left$Site)
```

```{r leftjk, echo = F}
knitr::kable(bat_left) |> 
  kableExtra::scroll_box(height = "300px") |> 
  kableExtra::kable_styling(full_width = F, html_font = 'Arial', font_size = 10) |> 
  kableExtra::column_spec(1:10, background = 'white', include_thead = T)
```

The left join is taking every row in the left data, `bat_sites`, and only the rows in the right data, `bat_cap`, that have a matching site. Note how site_004, which is only in the `bat_sites`, is included with NAs for the columns that came from the `bat_cap` data that didn't have a match. Site_006, which was only in the `bat_cap` data was dropped. 

<p class='note'><b>Coding tip:</b> I use left joins more than any other join because I'm usually joining tables that have a 1-to-many relationship, where the left dataset has 1 row for 1 or more rows in the right dataset. For example, say I have a dataset that only includes data for plots where an invasive species was detected and I want to do summary statistics that require the full number of plots. Using a left join, where the left dataset is a table of all of the plots and the right dataset is the invasive detections, will return the full set of plots to calculate summary statistics from. You may also have to fill 0s where NAs are introduced in the data before generating summary statistics, which should be done wisely.</p>

<p class='code'>Right join</p>
```{r rightj, fold.output = F}
bat_right <- right_join(bat_sites, bat_cap, by = "Site")
table(bat_right$Site)
```

```{r rightjk, echo = F}
knitr::kable(bat_right) |> 
  kableExtra::scroll_box(height = "300px") |> 
  kableExtra::kable_styling(full_width = F, html_font = 'Arial', font_size = 10) |> 
  kableExtra::column_spec(1:10, background = 'white', include_thead = T)
```

The right join is taking every row in the right data, `bat_cap`, and only the rows in the left data, `bat_sites`, that have a matching site. Note how Site_006, which is only in the `bat_cap`, is included with NAs for the columns that came from the `bat_sites` data that didn't have a match. Site_004, which was only in the `bat_sites` data was dropped. 

<p class = 'code'>Anti join to find sites not in bat_cap</p>
```{r aj, fold.output = F}
anti_join(bat_sites, bat_cap, by = "Site")
```

<p class = 'code'>Anti join to find sites not in bat_sites</p>
```{r aj2, fold.output = F}
anti_join(bat_cap, bat_sites, by = "Site")
```
</details>
<br>

<details open><summary class = 'drop'>Test your skills!</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Join the motile invertebrate count data frame to the motile invertebrate species table to get Invasive and Exotic columns added to the data. </strong>

<p class = 'code'>Import motinv data frames</p>
```{r challtr1}
# Read in motinv data if you haven't yet
motinv <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv")
head(motinv)

# Read in species table
motspp <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/motile_invert_species_table.csv")
head(motspp)

intersect(names(motinv), names(motspp)) # 3 columns in common
```

<details><summary class='drop2'>Answer</summary>
```{r challtr2a}
# left join species to motinv, because don't want to include species not found in count data
motinv_spp <- left_join(motinv, 
                        motspp, 
                        by = c("SpeciesCode", "ScientificName", "CommonName"))

head(motinv_spp)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find species in motspp data frame that don't have a match in the motinv data frame.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challtr2b}
# anti join of 
anti_join(motspp, motinv, by = c("SpeciesCode", "ScientificName", "CommonName"))

```
</details>
</div>

</details>
<br>
<details open><summary class='drop'>Rolling joins</summary>
There are a number of other more advanced joins out there, the rolling join being one of them. For more information on all possible joins, refer to <a href="https://r4ds.hadley.nz/joins.html">Chapter 19 in R for Data Science</a>.

Rolling joins can come in handy if the key values in your two datasets don't perfectly match, and you want to join on the closest match. An example of where I've used rolling joins is to relate timing of high tide to the nearest water temperature measurement from a HOBO logger. You can allow for the nearest match in both directions or specify the direction (e.g., => or <=). </li> 

<img src = "./images/rolling_join_R4DS.png" alt = 'rolling join' style = "display:block;margin-left:auto;margin-right:auto;width:300px;">
<center style = "font-size:11px;font-family:'Arial',sans-serif;"><a href="https://r4ds.hadley.nz/joins.html">Figure from R for Data Science</a></center>
<br>

Unfortunately, dplyr's rolling join doesn't perform the way I've needed it. It only matches in one direction, like the closest temperature measurement after high tide, or the closest temperature measurement before high tide. If you need to do a rolling join, the `data.table` package is your best bet. It requires learning a new syntax and coding approach, so I'm not covering it here. But it's helpful to know that if you're working with huge datasets, `data.table` tends to perform much faster than dplyr and may have more features for joining and summarizing your data than dplyr.

</details>
<br>
