#### Day 2 Questions
<details open><summary class = 'drop'>Load Data and Packages</summary>
If you're starting a new session to answer these questions, you'll need to load dplyr and read in the motile invertebrate data frame again.<br>

<p class = 'code'>Load dplyr</p>
```{r day2_ch, echo = F, include = F}
#------------------------------------
#     Day 2: Challenges Code 
#------------------------------------
```

```{r ldpd2c}
library(dplyr)
```

<p class = 'code'>Read in example motile invertebrate and point intercept data</p>
```{r readmotinv3bc1, eval = F}
# Point intercept data
pi_dat <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_Point_Intercept_data.csv")
# Motile invert count
motinv <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv")
# Motile invert site
motspp <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/motile_invert_species_table.csv")
# hobo temp data
temp_data <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/HOBO_temp_example.csv", skip = 1)[,1:3]
colnames(temp_data) <- c("index", "date_time", "tempF")
```

```{r motinvd2c1, echo = F}
motinv <- read.csv("./data/BASHAR_motile_invert_counts.csv")
pi_dat <- read.csv("./data/BASHAR_Point_Intercept_data.csv")
motspp <- read.csv("./data/motile_invert_species_table.csv")
```

</details>
<br>

<details open><summary class = 'drop'>Data Wrangling with dplyr </summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using `motinv`, how many species are found in PlotName = A1 in 2024?</strong>
<details><summary class='drop2'>Answer</summary>

```{r chal4dpc1}
# with brackets
A1_2024 <- motinv |> filter(PlotName == "A1" & Year == 2024)
nrow(A1_2024) # 3
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: What years have green crabs (Latin: Carcinus maenas) been detected? </strong>
<details><summary class='drop2'>Answer</summary>

```{r chal5bdpc1, fold.output = F}
gcrab <- motinv |> filter(ScientificName == "Carcinus maenas") |> 
  select(Year) |> unique()

gcrab
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find the highest value recorded in No.Damage column.</strong>
<details><summary class='drop2'>Answer</summary>
```{r maxnddpc1, fold.output = F}
max_nd <- max(motinv$No.Damage, na.rm = TRUE)
motinv |> filter(No.Damage == max_nd)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Fix the No.Damage typo by replacing 1960 with 196.</strong>

<details><summary class = 'drop2'>Answer</summary>
Let's say that you looked at the datasheet, and the actual count for No.Damage was 196 instead of 1960. You can change that value in the original CSV by hand. But even better is to document that change in code. There are multiple ways to do this. Two examples are below.

But first, it's good to create a new data frame when modifying the original data frame, so you can refer back to the original if needed. I also use a really specific filter to make sure I'm not accidentally changing other data.

<p class = 'code'>Replace 1960 with 196</p>
```{r trfxc1d2, fold.output = F}
# dplyr approach
motinv_fix <- motinv |> mutate(No.Damage = replace(No.Damage, No.Damage == 1960, 196))
range(motinv$No.Damage)
range(motinv_fix$No.Damage)

```
</details>
</div>
</details>

<br>

<details open><summary class = 'drop'>Conditionals</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using the motinv data, create a new column called trophic that indicates whether the species is an herbivore or predator. <br>
Hint: predator site codes are c("CARMAE", "NUCLAP"), and herbivore site codes are c("LITLIT", "LITOBT", "LITSAX", "TECTES"). </strong>
<details><summary class='code2'>Answer</summary>
```{r micondc1}
pred <- c("CARMAE", "NUCLAP")

# base R
motinv$trophic <- ifelse(motinv$SpeciesCode %in% pred, "predator", "herbivore")
table(motinv$trophic, motinv$SpeciesCode)

# tidyverse
motinv <- motinv |> mutate(trophic = ifelse(SpeciesCode %in% pred, "predator", "herbivore"))
table(motinv$trophic, motinv$SpeciesCode)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the motile invertebrate data, create a new column called count_level that has levels High, Medium, Low, based on No.Damage, where "High" is > 35, "Medium" is 10 - 35, and "Low" is < 10. </strong>
<details><summary class='code2'>Answer</summary>
```{r miabc1}
# Base R using a nested ifelse()
motinv$count_level <- 
  ifelse(motinv$No.Damage > 35, "High", 
         ifelse(motinv$No.Damage >= 10 & motinv$No.Damage <= 35, "Medium", "Low"))

table(motinv$count_level) # check that it worked
```

```{r miab2c1}
# Tidyverse using case_when() and between()
motinv <- motinv |> mutate(count_level = case_when(No.Damage > 35 ~ "High",
                                                   between(No.Damage, 10, 35) ~ "Medium",
                                                   No.Damage < 10 ~ "Low"))

table(motinv$count_level) # check that it worked
```
Note the use of the `between()` function that saves typing. This function matches as >= and <=. 
</details>
</div>
</details>
<br>

<details open><summary class='drop'>Summarizing</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using the point intercept data (`pi_dat`), calculate the average percent frequency of each non-vegetated substrate by year. Note that non-vegetated substrates are CoverCode = c('BOLT', 'ROCK', 'WATER'). </strong>
<details><summary class='code2'>Answer</summary>
```{r sumnvc1}
pi_nonveg <- pi_dat |> filter(CoverCode %in% c("BOLT", "ROCK", "WATER")) |> # filter nonveg grps
  summarize(avg_freq = mean(pct_freq), # calc avg.
            .by = c(SiteCode, Year, CoverCode, CoverType)) # grouping variables 

head(pi_nonveg) # check output
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the point intercept data (`pi_dat`), calculate the average percent frequency of each non-vegetated vs vegetated cover types by year. Note that non-vegetated substrates are CoverCode = c('BOLT', 'ROCK', 'WATER'). </strong>
<details><summary class='code2'>Answer</summary>
```{r sumnv2c1}
pi_subtype <- pi_dat |>
  mutate(sub_type = ifelse(CoverCode %in% c("BOLT", "ROCK", "WATER"), "nonveg", "veg")) |> # filter nonveg grps
  summarize(avg_freq = mean(pct_freq), # calc avg.
            .by = c(SiteCode, Year, sub_type)) |> # grouping variables 
  arrange(SiteCode, Year, sub_type) # sort variables

head(pi_subtype) # check output
```
</details>
</div>
</details>
<br>

<details open><summary class='drop'>Pivoting Tables</summary>

<div class="alert alert-info">
<strong>CHALLENGE: Use the `motinv_sum` data frame from the "Summarizing with dplyr" tab to pivot on SpeciesCode and mean_count, and fill the NAs with 0s. If you don't have the motinv_sum data frame handy, run the code below to create it. <br>
Hint: Drop the ScientificName and CommonName columns before you pivot.
</strong>

```{r motinvredo1c2}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))
```

<details><summary class='code2'>Answer</summary>
```{r pivwide2bc, fold.output = F}
motinv_wide <- motinv_sum |> 
  arrange(SpeciesCode) |> # sorting so columns are alphabetical 
  select(-ScientificName, -CommonName) |> 
  pivot_wider(names_from = SpeciesCode,
              values_from = mean_count, 
              values_fill = 0)

head(motinv_wide)

```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Use the `motinv_sum` data frame from the "Summarizing with dplyr" tab to pivot on Year and mean_count, fill the NAs with 0s, and add "yr_" to the column names to prevent column names starting with numbers. If you don't have the motinv_sum data frame handy, run the code below to create it. <br>
Hint: Drop the se_counts column before you pivot.
</strong>

```{r motinvredoc1}
# Fix the data issues again
motinv <- motinv |> 
  mutate(NoDamage_fix = replace(No.Damage, Damage == 1960, 196),
         Damage_fix = as.numeric(replace(Damage, Damage == "PM", NA)),
         total_count = NoDamage_fix + Damage_fix) |> 
  filter(QAQC == FALSE)

# Summarize the mean count per plot of each species by year and community type
motinv_sum <- motinv |> 
  summarize(mean_count = sum(total_count)/5, # 5 plots per site
            se_counts = sd(total_count)/sqrt(5), # 5 plots per site
            .by = c(SiteCode, Year, CommunityType, 
                    ScientificName, CommonName, SpeciesCode))
```

<details><summary class='code2'>Answer</summary>
```{r pivwide2c1c, fold.output = F}
motinv_wide_yr <- motinv_sum |> 
  arrange(Year) |> # sorting so columns are alphabetical 
  select(-se_counts) |> 
  pivot_wider(names_from = Year,
              values_from = mean_count, 
              values_fill = 0, 
              names_prefix = "yr_")

head(motinv_wide_yr)
```
</details>
</div>


<div class="alert alert-info">
<strong>CHALLENGE: Pivot the motinv_wide_yr data frame on the years columns, and remove the "yr_" from the year names using `names_prefix = 'yr_'`. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2a1c, fold.output = F}
motinv_long_yr <- pivot_longer(motinv_wide_yr, 
                               cols = -c(SiteCode, CommunityType, ScientificName, 
                                         CommonName, SpeciesCode),
                               names_to = "Year", 
                               values_to = "mean_counts", 
                               names_prefix = "yr_") # drops this string from values
```
</details>
</div>

</details><br>


<details open><summary class='drop'>Joining Tables</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Join the motile invertebrate count data frame to the motile invertebrate species table to get Invasive and Exotic columns added to the data. </strong>

<p class = 'code'>Import motinv data frames</p>
```{r challtr1cc}
# Read in motinv data if you haven't yet
motinv <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/BASHAR_motile_invert_counts.csv")
head(motinv)

# Read in species table
motspp <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/motile_invert_species_table.csv")
head(motspp)

intersect(names(motinv), names(motspp)) # 3 columns in common
```

<details><summary class='drop2'>Answer</summary>
```{r challtr2ac1}
# left join species to motinv, because don't want to include species not found in count data
motinv_spp <- left_join(motinv, 
                        motspp, 
                        by = c("SpeciesCode", "ScientificName", "CommonName"))

head(motinv_spp)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Find species in motspp data frame that don't have a match in the motinv data frame.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challtr2bc1c}
# anti join of 
anti_join(motspp, motinv, by = c("SpeciesCode", "ScientificName", "CommonName"))

```
</details>
</div>

</details><br>

<details open><summary class='drop'>Dates and Times</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you return date1 as YYYYMMDD (20260312)?</strong>
```{r}
# Create date1 if you don't have it already
date1 <- as.Date("3/12/2026", format = "%m/%d/%Y")
```
<details><summary class='drop2'>Answer</summary>
```{r chald31dc1}
format(date1, format = "%Y%m%d")
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you create a list of dates in 2026 that are evenly spaced by 3 months?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32bc1}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "3 months")
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you create a list of dates in 2026 that are evenly spaced by 1 week?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32cbc1}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "1 week")
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you extract the month as a number ranging from 1-12 in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatec1}
temp_data$month_num <- as.numeric(format(temp_data$timestamp, "%m"))
head(temp_data)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you extract the Julian date in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatejc1}
temp_data$julian <- as.numeric(format(temp_data$timestamp, "%j"))
head(temp_data)
```
</details>
</div>
</details><br>




