#### Day 3 Questions
<details open><summary class='drop'>Load Data and Packages</summary>
```{r day3_ch, echo = F, include = F}
#------------------------------------
#       Day 3: Challenges Code 
#------------------------------------
```

<p class = 'code'>Load packages and prep data for ggplot sections</p>
```{r loaddatc}
# packages
library(dplyr)
library(ggplot2)
library(patchwork) # for arranging ggplot objects
library(RColorBrewer) # for palettes
library(viridis) # for palettes

# load data
pcov <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/SHIHAR_photoplot_cover.csv") # import data

# define color and shape objects
cols <- c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
         "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
         "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")

shps <- c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
          "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25)

# Set x axis range
xrange <- range(pcov$Year)

```
</details><br>

<details open><summary class = 'drop'>Building a ggplot object</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, make the point size 1.5.</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga1c1}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 1.5) + # changed this line
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, make the error bars wider.</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga2c2}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 1.2) + # changed this line
  geom_point(color = "dimgrey", size = 2.5) + 
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") + 
  scale_shape_manual(values = shps, name = "Species Group") + 
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, change the x axis label to "Year".</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga3c1}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) + 
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = "Year", y = "Avg Percent Cover") + # changed this line
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

</details>
</div>
</details><br>

<details open><summary class = 'drop'>Improve formatting</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you change the smoother in the code below from LOESS to linear model and make the line dashed? Hint: method = 'lm'.</strong>
```{r gg12c1, results = 'hide'}
p6 + geom_smooth(se = F, span = 0.75)
```
<details><summary class='drop2'>Answer</summary>
```{r gga4bc1}
p6 + geom_smooth(method = 'lm', se = F, linetype = 'dashed')
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you make the legend title larger and bold, and legend text larger?</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga5bc1}
p6 + theme(legend.title = element_text(size = 12, face = 'bold'),
           legend.text = element_text(size = 11))
```
</details>
</div>
</details><br>

<details open><summary class = 'drop'>Palettes</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you specify the 'RdYlBu' palette instead of the ones used above?</strong><br>
Hint: Start with p_pal to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpalc1, fig.height = 4.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Species Group", palette = "RdYlBu", 
                            aesthetics = c("fill", "color"))  
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Create your own palette with at least three colors.</strong><br>
Hint: Start with p_heat to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal2dc1, fig.height = 4.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "#3E693D", mid = "#FDFFC7", high = "#7A6646", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(chem$Temp_F), 
                               name = "Temp. (F)") 
```
</details>
</div>

</details><br>
