---
output: 
  html_document:
    css: custom_styles.css
---

```{r day2_gg, echo = F, include = F}
#------------------------------------
#    Day 2: Intro to ggplot Code 
#------------------------------------
```

```{r optchunk, echo = F}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align = 'center', fig.height = 4, fig.width = 6)
```

#### Intro to `ggplot2`
<details open><summary class='drop'>Intro to `ggplot2`</summary>
The `ggplot2` package is the most popular `R` package for plotting. It takes a little effort to learn how the pieces of a ggplot object fit together. However, once you get the hang of it, you can create and customize a large variety of attractive plots with just a few lines of `R` code. The package is called `ggplot2` because originally there was `ggplot`. The developer, Hadley Wickham, didn't want to break the original package to improve the package, so created ggplot2.  

[The ggplot2 online book]("https://ggplot2-book.org/") and [cheatsheets](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf") can be very helpful while you are learning to use the `ggplot2` package.

The `ggplot2` package was developed using the grammar of graphics as the underlying philosophy, which basically breaks a plot up into individual building blocks related to aesthetics (e.g., color, size, shape), geometries (e.g. points, lines, boxes), and themes (e.g. axis label font size, legend placement, etc.).

Important concepts with `ggplot2`:
<ul>
<li><b>ggplot(data, aes()):</b> Every ggplot object starts with this line, which tells R the data you're plotting, and which variables you're plotting where in `aes()` argument.</li>
<li><b>data: </b> every plot requires data. The first argument for every `ggplot()`call is the data. This also means you can pipe data into a ggplot object. </li>
<li><b>aes: </b> Short for aesthetics. This is where you tell ggplot what your x and y variables are. If you want aesthetics, like color, fill, or size to vary by the data (e.g., color code a figure by park, use a dashed vs. solid line to distinguish between significant/non-significant trend, etc.), those variables are specified within the `aes()` argument either at the `ggplot()` level, or within the specific geom. </li>
<li><b>geom: </b> Short for geometries, geoms represent what you see in the plot, such as the points in a scatter plot, the lines in a trend plot, or the boxes in a boxplot.</li>
<li><b>scale: </b>Scales go hand in hand with `aes()`. If you specify `aes(color = park)`, then scale is where you can specify a custom color for each park instead of ggplot's default color scheme. The scale is where you can set the labels of groups in a legend (if different than how the data are labeled). You can also customize axis ranges, breaks, and labels with different scales. </li>
<li><b>theme: </b>This is where you can change the format of the figure, such as removing the gridlines in the default ggplot format, making axis labels larger, changing the position of the legend, etc.</li>
<li><b>facet: </b>Faceting the data allows you to graph multiple plots based on a grouping variable (e.g., site, species, year, etc.) in the same ggplot object.</li>
</ul>

</details>
<br>

#### Building a ggplot 
<details open><summary class = "drop">Building up a ggplot</summary>
We're going to use the summarized photo plot cover to create the plot below, and will work through the code one piece at a time. 

```{r echo = F, fig.height = 5, fig.width = 7, fold.plot = F}
library(ggplot2) # load ggplot
pcov <- read.csv("./data/SHIHAR_photoplot_cover.csv") # import data

cols <- c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
         "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
         "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")

shps <- c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
          "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25)

xrange <- range(pcov$Year)

ggplot(data = pcov, aes(x = Year, y = median_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  #geom_smooth(se = F, span = 0.75) +
  #geom_line(linewidth = 0.6) +
  geom_errorbar(aes(ymin = q25_cover, ymax = q75_cover), linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```


<p class = 'code'>Import photoplot cover data from Ship Harbor</p>
```{r gg1, eval = F}
library(ggplot2) # load package

pcov <- read.csv("https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/SHIHAR_photoplot_cover.csv") # import the data set

head(pcov) # check out data set
```

<p class = 'code'>Create the ggplot template of median cover over time.</p>
I'm going to assign this to an object named `p`. We're going to have unique colors and shapes for the CoverCode column in our data, so we need to indicate that in the `aes()` along with our x and y variables. If we didn't include color, fill, and shape in the `aes()`, the points would all be the same color (default = black) and shape (default is filled circle)
```{r gg2, fold.plot = F}
p <- ggplot(data = pcov, aes(x = Year, y = median_cover, 
                             color = CoverCode, 
                             fill = CoverCode,
                             shape = CoverCode))

p
```

<p class = 'code'>Add point and errorbar geometry</p>
The order of geoms is the order they're drawn. I generally prefer the look of the points after the error bars. 
```{r gg3, fold.plot = F}
p2 <- p + 
  geom_errorbar(aes(ymin = q25_cover, ymax = q75_cover)) +
  geom_point(color = 'dimgrey', size = 2.5) # setting point outline to dark grey (bear with me)

p2
```

The default colors and symbols in ggplot aren't great. We're going to start by specifying our own colors and shapes manually. Then we'll use color palettes from different packages. Specifying shapes in R requires knowing the shape's symbol code. To view that, run `?points`, or search "pch in R plot" and you'll get the info below. Note that 0-14 are just lines with no fill. To change their color, use the color aesthetic. Symbols 15-20 are solid, but also use color to change their aesthetic. Symbols 21-25 have both a color (outline) and fill (inside) aesthetic.

<center><img src = "./images/r_pch_symbols.png" alt = 'R pch symbols' style = "width:250px;background:white;"><br>
<p class = "cap">Figure of symbol codes in R.</p></center>
<br>

<p class = 'code'>Specify colors and shapes</p>
```{r gg4, fold.plot = F}
p3 <- p2 + 
  scale_fill_manual(values = c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
                               "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
                               "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")) +
  scale_color_manual(values = c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
                               "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
                               "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")) +
  scale_shape_manual(values = c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
                                "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25))
p3
```
The code above was redundant and tedious to code. Changing colors means changing them in two places. Imagine you're making one of these plots for every rocky intertidal site in Acadia. It quickly becomes cumbersome. A more efficient way to code this is below. 

<p class = 'code'>Assign colors and shapes to variables </p>
Adding the name to the scales renames the legend title. But to keep the shapes and colors in the same legend, you have to name all of them the same name.
```{r gg4b, fold.plot = F}
cols <- c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
         "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
         "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")

shps <- c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
          "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25)

p3 <- p2 + 
  scale_fill_manual(values = cols, name = "Species Group") +
  scale_color_manual(values = cols, name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group")
  
p3
```

<p class = 'code'>Improve axes ticks and labels </p>
```{r gg5, fold.plot = F}
# Determine year range (so not hard coded/easily updated in future)
xrange <- range(pcov$Year)

p4 <- p3 + 
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") 
  
p4
```

<p class = 'code'>Improve theme components</p>
```{r gg6, fold.plot = F} 
p5 <- p4 +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), # angle x text
          panel.grid.major = element_blank(), # turns of major grids
          panel.grid.minor = element_blank(), # turns off minor grids
          panel.background = element_rect(fill = 'white', color = 'dimgrey'),# makes background white 
          legend.key = element_blank()) # removes square fill around symbols in legend

p5
```

<p class = 'code'>Facet on CommunityType</p>
```{r gg7, fold.plot = F, fig.height = 5, fig.width = 7,}
p6 <- p5 + facet_wrap(~CommunityType)

p6  

```

<p class = 'code'>Full code with code to save figure</p>
```{r gg8, fold.plot = F, fig.height = 5, fig.width = 7,}
ggplot(data = pcov, aes(x = Year, y = median_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = q25_cover, ymax = q75_cover), linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```
```{r ggs, eval = F}
ggsave("SHIHAR_photoplot_cover.png", height = 8, width = 7)
```

</details>
<br>

#### Other geometries
<details open><summary class = 'drop'>Variations on figures</summary>
<p class = 'code'>Add line geometry to figure</p>
```{r gg11, fold.plot = F}
p6 + geom_line(linewidth = 0.8)
```

<p class = 'code'>Add smoother to figure</p>
```{r gg12, fold.plot = F}
p6 + geom_smooth(se = F, span = 0.75)
```

<p class = 'code'>Move legend to the bottom</p>
```{r gg13, fold.plot = F}
p6 + theme(legend.position = 'bottom')
```

<p class = 'code'>Stacked bars instead of points </p>
```{r gg9, fold.plot = F}
ggplot(data = pcov, aes(x = Year, y = median_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_bar(stat = 'identity', position = 'fill', color = 'dimgrey') +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)



```

<p class = 'code'>Bars with error bars instead of points (note I filtered on Barnacle CommunityType) </p>
```{r gg10, fold.plot = F}
ggplot(data = pcov |> filter(CommunityType == "Barnacle"), # note filter 
       aes(x = Year, y = median_cover, 
           color = CoverCode, group = CoverCode,
           fill = CoverCode, shape = CoverCode)) +
  geom_bar(stat = 'identity', position = 'dodge', color = 'dimgrey') + # new line
  geom_errorbar(aes(ymin = q25_cover, ymax = q75_cover), linewidth = 0.6) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CoverCode) # Different facet


```

</details><br>



