---
output: 
  html_document:
    css: custom_styles.css
---

#### Customizing ggplot
<details open><summary class = 'drop'>Load and prep data</summary>
For this section, we're going to use NETN water quality data to customize ggplot objects. This is an abbreviated dataset from the NETN water package. The data contains surface lake measurements recorded using a YSI in a subset of lakes in Acadia NP. We'll filter the data to plot at different combinations of parameters and sites. 

```{r day3_gg, echo = F, include = F}
#------------------------------------
#       Day 2: Adv. ggplot Code 
#------------------------------------
```

```{r chem, eval = F}
library(dplyr)
library(ggplot2)
library(patchwork) # for arranging ggplot objects
library(RColorBrewer) # for palettes
library(viridis) # for palettes
chem <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_water_chemistry_data.csv")
str(chem)
```
```{r chem2, echo = F}
library(viridis)
library(RColorBrewer)
library(scales)
library(viridis)
chem <- read.csv("./data/NETN_water_chemistry_data.csv")
str(chem)
```

To start with, we're going to plot temperature data for 8 lakes monitored in Acadia NP. Before we start plotting, we need to convert the EventDate column to a date type, and will extract a year, month, and day of year column for easier plotting later on. 

<p class ='code'>Add date columns to chem, then filter on sites and Temp_F. </p>
```{r chem3}
chem <- chem |> mutate(date = as.Date(EventDate, "%m/%d/%Y"),
                       year = as.numeric(format(date, "%Y")),
                       mon = as.numeric(format(date, "%m")),
                       doy = as.numeric(format(date, "%j"))) 

ACAD_lakes <- c("ACBUBL", "ACEAGL", "ACECHO", "ACJORD", 
                "ACLONG", "ACSEAL", "ACUHAD", "ACWHOL")

lakes_temp <- chem |> filter(SiteCode %in% ACAD_lakes) |> 
  filter(Parameter %in% "Temp_F") 

head(lakes_temp)
```
</details>
<br>
<details open><summary class = 'drop'>Customizing shapes and colors manually</summary>
Now that we have the data set up, we're going to make a line and point time series plot of temperature for the different sites.

<p class = 'code'>Make a generic plot with the black and white built in theme</p>
```{r plot1, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value)) + 
  theme_bw() +
  geom_point() 
```

<p class = 'code'>Set color and symbol by SiteCode using default colors and shapes</p>
```{r plot2, warning = T, fold.plot = F}
ggplot(lakes_temp, 
       aes(x = date, y = Value, color = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point() 

```
Note the warning that we have 8 groups, but ggplot by default only provides 6 different shapes. The 220 rows removed corresponds to the points that were dropped that belonged to the 7th and 8th sites (ACUHAD and ACWHOL). To use 8 symbols, you have to specify them manually, which we'll do next.

In addition, the default colors in ggplot aren't great. Whenever you see these colors used in publications, you kind of know the author either barely knows ggplot or is lazy. We're going to start by specifying our own colors and shapes manually. Then we'll use color palettes from different packages. 

Before you start plotting, it's helpful to know what point symbol codes are. To view that, run `?points`, or search "pch in R plot" and you'll get the info below. Note that 0-14 are just lines with no fill. To change their color, use the color aesthetic. Symbols 15-20 are solid, but also use color to change their aesthetic. Symbols 21-25 have both a color (outline) and fill (inside) aesthetic.

<center><img src = "./images/r_pch_symbols.png" alt = 'R pch symbols' style = "width:250px;background:white;"><br>
<p class = "cap">Figure of symbol codes in R.</p></center>
<br>

<p class='code'>Specify manual color and shape connected to SiteCode</p>
```{r plot3, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, 
                       shape = SiteCode)) + 
  theme_bw() +
  geom_point() +
  scale_color_manual(values = c("ACBUBL" = "#7FC7E1", "ACEAGL" = "#ED5C5C", 
                                "ACECHO" = "#F59617", "ACJORD" = "#A26CF5", 
                                "ACLONG" = "#1952CF", "ACSEAL" = "#F3F56C", 
                                "ACUHAD" = "#FFA6C8", "ACWHOL" = "#8FD184")) +
  scale_fill_manual(values = c("ACBUBL" = "#7FC7E1", "ACEAGL" = "#ED5C5C", 
                               "ACECHO" = "#F59617", "ACJORD" = "#A26CF5", 
                               "ACLONG" = "#1952CF", "ACSEAL" = "#F3F56C", 
                               "ACUHAD" = "#FFA6C8", "ACWHOL" = "#8FD184")) +
  scale_shape_manual(values = c("ACBUBL" = 21, "ACEAGL" = 23, 
                                "ACECHO" = 24, "ACJORD" = 25, 
                                "ACLONG" = 23, "ACSEAL" = 21, 
                                "ACUHAD" = 25, "ACWHOL" = 24))

```

The code above gave us a different color/fill and shape for each point, but coding it was cumbersome. Specifying the colors like above gets tedious fast. Imagine needing to make a plot for a dozen different parameters. A more efficient approach is defining them outside of ggplot, and then referencing them in the plot, like the example below.

<p class='code'>Specify manual color and shape more efficiently</p>
```{r plot4, fold.plot = F}
site_cols <- c("ACBUBL" = "#7FC7E1", "ACEAGL" = "#ED5C5C", 
               "ACECHO" = "#F59617", "ACJORD" = "#A26CF5", 
               "ACLONG" = "#1952CF", "ACSEAL" = "#F3F56C", 
               "ACUHAD" = "#FFA6C8", "ACWHOL" = "#8FD184")
site_shps <- c("ACBUBL" = 21, "ACEAGL" = 23, 
               "ACECHO" = 24, "ACJORD" = 25, 
               "ACLONG" = 23, "ACSEAL" = 21, 
               "ACUHAD" = 25, "ACWHOL" = 24)

ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point() +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") 

```

<p class='code'>Similar plot, but make the outline the same color for each point and increase point size</p>
```{r, plot5, warning = T, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point(color = "dimgrey", size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") 
```

The warning is there because you have color in the `aes()` and in the `geom_point()`, but it's not a problem. The ggplot2 package is pretty chatty in the console. It's good to read the warnings to make sure you didn't drop values you wanted to plot (e.g. ACWHOL dropped in previous example), but often they're not issues.

<div class="alert alert-info">
<strong>CHALLENGE: Review labels from Day 2. How would you change the x-axis label to "Year", and the y-axis label to "Temp. (F)"?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3lab}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point(color = "dimgrey", size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(x = "Year", y = "Temp. (F)") # answer
```
</details>
</div>

</details><br>

<details open><summary class = 'drop'>Add lines and smoothers</summary>
Now we're going to play with adding lines to the graphs. First we'll add the `geom_line()` to see how it looks. Notice the order has the line plotting before the point, so it doesn't cross over the points. 
```{r plot6l, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_line() +
  geom_point(color = "dimgrey") +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake")  +
  labs(x = "Year", y = "Temp. (F)") 
```

It's not easy to follow the line and not really what we're looking for at this scale of the data (2006-2025). This is where we have to think about what we're actually interested in, and in this case, it's whether temperature is changing over time. This is where the `geom_smooth()` is really helpful. The `geom_smooth()` plots a line assuming the y ~ x formula (unless you specify a different formula). By default the method is a LOESS smoother, but you can specify a range of methods, including linear regression by adding `method = 'lm'` to `geom_smooth()`. 

Note that I turned off the standard error ribbon that plots by default using `se = FALSE`. It's too busy for this plot. I also don't use the SE unless I've fit an actual model and checked the diagnostics. The status under the hood of `geom_smooth()` are also pretty black boxy, and I don't always know if I can trust its calculation of SE. 

I added a transparency via `alpha = 0.5` to the `geom_point()`, so the lines show up better. 

<p class='code'>Add a LOESS smoother and make points more transparent</p>
```{r plot6, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = FALSE, span = 0.5, linewidth = 1) +
  geom_point(color = "dimgrey", alpha = 0.5) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake")  +
  labs(x = "Year", y = "Temp. (F)") 
```

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the smooth from LOESS to linear?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3lm}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_point(color = "dimgrey", alpha = 0.5) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake")  +
  labs(x = "Year", y = "Temp. (F)") 
```
</details>
</div>

</details><br>

<details open><summary class = 'drop'>Facets and grids</summary>
<details open><summary class = 'drop2'>Facets</summary>
The plot looks okay so far, but there are so many points, it's hard to see what's going on. This is where facets are useful. If your data have grouping variables, in this case SiteCode, then you can plot separate panels for each of the grouping levels. The code below plots each site separately. I used `ncol = 4` to set the number of columns that result from the facet wrap. 

<p class='code'>Facet on SiteCode</p>
```{r plot7, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_site <- 
  ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                         fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temperature (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 4)
p_site
```

Faceting on year can also be a handy way to see how consistent seasonal patterns are across years. Note that we changed the x variable from date to doy (day of year). I also filtered on the dataset within the ggplot line to only include years after 2015, increased the point size, and switched to `geom_line()` instead of the smoother to just connect the points. We'll revisit this plot to use more meaningful x-axis labels in the next section. 

<p class = 'code'>Facet on year instead of site</p>
```{r plotyr, fold.output = F, fig.height = 3.5, fig.width = 7, fold.plot = F}
p_year <- 
  ggplot(lakes_temp |> filter(year > 2015), 
         aes(x = doy, y = Value, color = SiteCode, 
             fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_line(linewidth = 0.7) +
  geom_point(color = "dimgrey", size = 2.5) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temperature (F)", x = "Year") +
  facet_wrap(~year, ncol = 3) 
p_year
```

<div class="alert alert-info">
<strong>CHALLENGE: Recreate the plot below</strong>
Note that the symbol outline is black. The alpha level is 0.6.
```{r challd3copy, echo = F, fold.plot = F, fig.height = 7, fig.width=5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')
```

<details><summary class='drop2'>Answer</summary>
```{r challd3copya, fig.height = 7, fig.width = 5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')

```

</details>
</div>

</details><br>

<details open><summary class = 'drop2'>Grids in patchwork</summary>
Faceting is helpful when your observations are all within the same column. But say you have data in multiple columns (e.g., each water quality parameter is a column) and want to arrange those plots into a grid. Faceting won't help because the data to plot are in different columns. There are multiple packages that make it easy to arrange multiple plots into a grid to look similar to faceted plots. Packages include `grid` (and `gridExtra`), `cowplot`, `ggpubr`, and `patchwork`. We're going to use `patchwork`, a relative newcomer, and one of the easiest I've found to code and customize. Here we're going to plot pH, temperature, DO, and conductance for Jordan Pond and arrange them using patchwork.

The patchwork package has a lot of options to customize plot layouts. See the <a href="https://patchwork.data-imaginist.com/">patchwork</a> package website for more information.

<p class = 'code'>Prepare the data to plot</p>
```{r}
pH <- chem |> filter(SiteCode == "ACJORD") |> filter(Parameter == "pH")
temp <- chem |> filter(SiteCode == "ACJORD") |> filter(Parameter == "Temp_F")
dosat <- chem |> filter(SiteCode == "ACJORD") |> filter(Parameter == "DOsat_pct")
cond <- chem |> filter(SiteCode == "ACJORD") |> filter(Parameter == "SpCond_uScm")

p_pH <-
  ggplot(pH, aes(x = date, y = Value)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "pH", x = "Year")  

p_temp <-
  ggplot(temp, aes(x = date, y = Value)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "Temp (F)", x = "Year")  

p_do <-
  ggplot(dosat, aes(x = date, y = Value)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "DO (%sat.)", x = "Year")  

p_cond <-
  ggplot(cond, aes(x = date, y = Value)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "Spec. Cond. (uScm)", x = "Year")  

```

<p class='code'>Arrange plots using patchwork</p>
This is almost too easy to be true, but it really is this easy with patchwork. The patchwork package includes a bunch of options to customize sizes, add annotation, sharing axes across plots. 
```{r patch1, fold.plot = F, fig.height = 4, fig.width = 5}
library(patchwork)
p_pH + p_temp + p_do + p_cond
```

<p class='code'>Arrange plots using patchwork in column of 4 and share x axis. </p>
You can also collect the legend using a similar approach to collecting the axes.
```{r patch2, fold.plot = F, fig.height = 6, fig.width = 3}
library(patchwork)
p_pH / p_temp / p_do / p_cond + plot_layout(axes = "collect_x")
```

</details>
<br>

<details open><summary class = 'drop'>Axes for dates</summary>
Starting with the plot faceted on SiteCode, I don't like that the first year in the data (2006) is missing from the axis. The x-axis is a Date type, which gives us some useful options to set breaks and labels. Below, I set the breaks to be every 2 years and to label only years (`%Y`). I added the `theme()` for the `axis.text.x` to make the years plot vertically, and the `vjust = 0.5` centers the labels on the tick marks. Note that I'm assigning the code below to the object `p`, so I don't have to keep typing the original ggplot code over and over.

<p class = 'code'>Improve x-axis labels</p>
```{r plot8, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_site2 <- p_site + 
  scale_x_date(date_labels = "%Y", date_breaks = "2 years") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

p_site2
```

Returning to the `p_year` plot faceted by year. Instead of the day of year on the x-axis, we want to manually set up date axis labels that include the month and day at the beginning of each month. This is a bit trickier, because `doy` isn't a Date type. It's still doable, there may be easier ways to do this. 

<p class = 'code'>Manually set up date labels for doy axis</p>
```{r plotyr2, fold.output = F}
# Find range of months in the data
range_mon <- range(lakes_temp$mon)
range_mon # 5:10

# Set up the date range as a Date type
range_date <- as.Date(c("5/1/2025", "11/01/2025"), format = "%m/%d/%Y")
axis_dates <- seq.Date(range_date[1], range_date[2], by = "1 month")
axis_dates
axis_dates_label <- format(axis_dates, "%b-%d")

# Find the doy value that matches each of the axis dates
axis_doy <- as.numeric(format(axis_dates, "%j"))
axis_doy
axis_doy_limits <- c(min(axis_doy)-1, max(axis_doy) + 1)
# Set the limits of the x axis as before and after the last sample, 
# otherwise cuts off May 1 in axis
```

```{r yearplot, fig.height = 3.5, fig.width = 7, fold.plot = F}
p_year + 
  scale_x_continuous(limits = axis_doy_limits,
                     breaks = axis_doy, 
                     labels = axis_dates_label) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the plot called p_site2 so that x-axis labels were every 4 years, and angle is 45 degrees instead of 90?</strong><br>
Hint: start with p_site, so you don't have to write that much code. You'll also need to tweak vjust and hjust to make the labels line up properly.
<details><summary class='drop2'>Answer</summary>
```{r challyears, fig.height = 4.5, fig.width = 7}
p_site + 
  scale_x_date(date_labels = "%Y", date_breaks = "4 years") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
</details>
</div>

</details>
<br>

<details open><summary class = 'drop'>Legends</summary>
One of the easier tasks to do with legends is to change the location of it. In the next plot we'll put the legend on the bottom. If you don't want to plot a legend (don't really need it when lakes are faceted, for example, you can turn the legend off using `legend.position = 'none'`). 

<p class = 'code'>Move legend to bottom</p>
```{r plot8b, fig.height = 4.5, fig.width = 7, fold.plot = F}
p_site3 <- p_site2 + theme(legend.position = 'bottom')
p_site3
```

The plots above made legends seem easy, but legends in ggplot can be really tedious. For example, the legend only shows up if you're setting grouping variables in the `aes()`. There was no legend in the very first plot we made because everything was the same color and symbol. If, for example, you want to plot thresholds on a plot, their color needs to be added to the `scale_color_manual()` to show in the legend. Let's pretend, for example, that 50F and 75F are lower and upper water quality thresholds that we want to plot.

<p class = 'code'>Add horizontal threshold lines to the plot</p>
```{r plot9la, fig.height = 4.5, fig.width = 7, fold.plot = F}
p_site3 + geom_hline(yintercept = 75, linetype = 'dashed', linewidth = 1) +
          geom_hline(yintercept = 50, linetype = 'dotted', linewidth = 1)
```

These lines are not showing in the legend. To make them show, you need to wrap them in `aes()`, as below. Note the difference in how linetype is specified between the above example, where we are specifying whether the line is dashed or dotted. Inside the `aes()`, `linetype` is being used to label the line in the legend. We then have to use `scale_linetype_manual()` to indicate the type of line to plot. Admittedly, it took me a couple of Stackoverflow posts to figure out how to make this work properly. This stuff can be tedious.

<p class = 'code'>Add horizontal threshold lines to the plot and legend</p>
```{r plot9l, fig.height = 4.5, fig.width = 7, fold.plot = F}
p_site4 <- p_site3 + 
  geom_hline(aes(yintercept = 75, linetype = "Upper"), linewidth = 1) +
  geom_hline(aes(yintercept = 50, linetype = "Lower"), linewidth = 1) +
  scale_linetype_manual(values = c("dashed", "dotted"), name = "WQ Threshold")

p_site4
```

Another option is turning certain geometries off in the legend via `show.legend = F`. Let's say we don't like the lines in the legend. I have to go back to the full code to change the geom legend settings. I'm also overriding the alpha level of the symbols, so they show up better in the legend.

<p class = 'code'>Remove smoothed lines from legend and increase alpha of symbols in legend</p>
```{r plot11b, fig.height = 4.5, fig.width = 7, fold.plot = F}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5, show.legend = FALSE) + 
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) + 
  
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temperature (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 4) + 
  scale_x_date(date_labels = "%Y", date_breaks = "2 years") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), 
        legend.position = 'bottom') + 
  
  geom_hline(aes(yintercept = 75, linetype = "Upper"), linewidth = 1) +
  geom_hline(aes(yintercept = 50, linetype = "Lower"), linewidth = 1) +
  scale_linetype_manual(values = c("dashed", "dotted"), name = "WQ Threshold") +
  
  guides(fill = guide_legend(override.aes = list(alpha = 1)))

```

Another option is to remove the SiteCode fill, color, and shape from the legend via `guides()`. You can also do this by turning a geom off using `show.legend = F` like we did with the smoother. In the code below we are telling ggplot that any shape, color, or fill used in an `aes()` should not be included in the legend. The WQ thresholds still show up because their aes was linetype. 

Finally, the width of the lines in the WQ thresholds makes it appear that the Lower threshold is solid instead of dashed. I'll use an option in theme to make the key wider.

<p class = 'code'>Remove SiteCode keys from legend and increase key width of lines.</p>
```{r plot9ld, fig.height = 4, fig.width = 7, fold.plot = F}
p_site4 + guides(shape = 'none', color = 'none', fill = 'none') +
          theme(legend.key.width = unit(0.8, "cm"))

```
</details>
<br>

#### ggplot Palettes

<details open><summary class = 'drop'>External ggplot Palettes</summary>
A number of R packages have color palettes available that are colorblind friendly. The two most commonly used are RColorBrewer and viridis. Both packages include palettes that are for three main types of data: 
 <ul>
 <li><b>Sequential</b>: for continuous variables that grade from low to high or vice versa. These tend to be one hue that increases in saturation as values increase.</li>
 <li><b>Diverging</b>: for data where low values have different colors than high values (e.g. starts red, then grades to blue).</li>
 <li><b>Qualitative</b>: data for categorical variables where all colors are a similar saturation but different hues.</li>
 </ul>

<details open><summary class='drop2'>RColorBrewer palette</summary>
The palettes in RColorBrewer can be viewed by running the code below. The first group shows the sequential palettes (e.g. YlOrRd - Yellow Orange Red). The second group shows the qualitative colors. The last group shows the diverging palettes. The main drawback of these palettes is they are limited by the number of levels in your data. So, if you specify `Set2` to color code different levels of a factor, there are only 8 colors available to you. If your factor has more than 8 levels (e.g., 9 sites, 10 parks, etc.), then the levels beyond 8 won't get plotted and you'll get a warning in the console similar to what we saw for ggplot's default number of symbols.

```{r day3_gp, echo = F, include = F}
#------------------------------------
#     Day 3: ggplot Palettes Code 
#------------------------------------
```

<p class = 'code'>View RColorBrewer palettes</p>
```{r rbcolpal_code, eval = F}
display.brewer.all(colorblindFriendly = TRUE)
```
<center><img src = "./images/RColorBrewer_palettes.png" alt = 'RColorBrewer palettes' style = "width:350px;"></center>
<br>

Going back to the temperature plots we made before, we'll use RColorBrewer to color code each site instead of doing this manually. We'll build the plot in the next chunk, that we then change the color palettes with in later plots.

<br>
<p class = 'code'>Create basic plot</p>
```{r tempp2}
p_pal <- ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                                fill = SiteCode, shape = SiteCode)) + 
         theme_bw() +
         geom_smooth(se = F, span = 0.5, linewidth = 1) + 
         geom_point(color = "dimgrey", alpha = 0.5, size = 2) + 
  
         scale_shape_manual(values = site_shps, name = "Lake") +
         labs(y = "Temperature (F)", x = "Year") +
  
         facet_wrap(~SiteCode, ncol = 4) + 
  
         scale_x_date(date_labels = "%Y", date_breaks = "2 years") +
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) + 
  
         geom_hline(aes(yintercept = 75, linetype = "Upper"), linewidth = 1) +
         geom_hline(aes(yintercept = 50, linetype = "Lower"), linewidth = 1) +
         scale_linetype_manual(values = c("dashed", "dotted"), name = "WQ Threshold")
```

<br>

<p class = 'code'>Use Set2 palette on temperature plot and remove transparency of symbols in legend</p>
```{r set2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Lake", palette = "Set2", aesthetics = c("fill", "color"))  +
         guides(fill = guide_legend(override.aes = list(alpha = 1))) # solid symbols in legend

```

Note how I used the aesthetics in the `scale_color_brewer()` to set fill and color as the same time. We could have done this in the code above too. I also changed the symbols in the legend to not be transparent, so they're easier to see using the `override.aes` in the guide.

<br>

<p class = 'code'>Use Dark2 palette on temperature plot</p>
```{r colbred2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Lake", palette = "Dark2", aesthetics = c("fill", "color"))  +
         guides(fill = guide_legend(override.aes = list(alpha = 1))) # makes symbols not transparent in legend


```
</details>
<br>

<div class="alert alert-info">
<strong>CHALLENGE: How would you specify the 'RdYlBu' palette instead of the ones used above?</strong><br>
Hint: Start with p_pal to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal, fig.height = 4.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Lake", palette = "RdYlBu", aesthetics = c("fill", "color"))  +
         guides(fill = guide_legend(override.aes = list(alpha = 1))) # makes symbols not transparent in legend
```
</details>
</div>

<details open><summary class = 'drop2'>viridis palettes</summary>
The viridis package comes with 8 palettes. The benefit of viridis is the number of levels is not limited to 8 like RColorBrewer. The palette options are below for 12 levels. 

<center><img src = "./images/viridis/viridis-scales.png" alt = 'viridis palettes' style = "width:450px;"></center>

<br>

<p class = 'code'>View viridis palettes with hexcodes</p>
You can view the hexcodes of the different palettes by running the code below. Just change `viridis()` to one of the other palette names to get the hexcodes for those levels. 
```{r vir, eval = F}
# viridis 
scales::show_col(viridis(12), cex_label = 0.45, ncol = 6)
```
<center><img src = "./images/viridis/viridis.png" alt = 'viridis palette' style = "width:350px;"></center>
<br>
<p class = 'code'>Use viridis default palette on temperature plot</p>
The `scale_color_viridis_d()` selects the viridis palette option (purple, green, yellow) for discrete values (i.e. categories). For a continuous scale (e.g. temperature), you would specify `scale_color_viridis_c()`. 
```{r vir2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal + scale_color_viridis_d(name = "Lake", aesthetics = c("fill", "color"))  #default viridis 
```

<p class = 'code'>Use turbo palette on temperature plot</p>
The `scale_color_viridis_d()` selects the viridis palette option (purple, green, yellow) for discrete values (i.e. categories). For a continuous scale (e.g. temperature), you would specify `scale_color_viridis_c()`. 
```{r vir2b, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal + scale_color_viridis_d(name = "Lake", aesthetics = c("fill", "color"), option = 'turbo') 
```
</details>
<br>
<details open><summary class = 'drop2'>Continuous palette with heatmaps</summary>
Heatmaps via `geom_tile()` are a place where viridis palettes are especially helpful producing useful sequential or diverging color palettes. We'll use the temperature data to plot heatmaps by month for each site. Heatmaps are a bit different than other plots we've seen, as the x and y values create a discrete grid, and the color in the cell represents the value for that level of x and y. That means we have to change how the x, y and color aesthetics are specified. Here we will plot temperature by month and year faceted on site.

<p class = 'code'>Basic heatmap code </p>
Note the use of base R's `month.abb` to set the labels on the x-axis. The `month.abb` is a vector of the 12 months abbreviated as 3 letters. By setting 5:10, I'm taking the months May - Oct.
```{r virheat}
p_heat <- 
ggplot(lakes_temp, aes(x = mon, y = year, color = Value, fill = Value)) + 
  theme_bw() +
  geom_tile() + 
  labs(y = "Year", x = "Month") +
  facet_wrap(~SiteCode, ncol = 4) + 
  scale_x_continuous(breaks = c(5, 6, 7, 8, 9, 10),
                     limits = c(4, 11), 
                     labels = month.abb[5:10]) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

<p class = 'code'>Plot heatmap with viridis continuous palette </p>
```{r virheat1, fold.output = F, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_viridis_c(name = "Temp. (F)", aesthetics = c("fill", "color")) 
```

<p class = 'code'>Plot heatmap with plasma continuous palette, reverse scale </p>
```{r virheatp, fold.output = F, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_viridis_c(name = "Temp. (F)", aesthetics = c("fill", "color"), 
                               option = "plasma", direction = -1) 
```

</details>
</details>
<br>
<details open><summary class = 'drop'>Create your own color ramp</summary>
You can also create your own color ramp via `scale_color_gradient()`, which creates a 2-color gradient,  `scale_color_gradient2()`, which creates a diverging color gradient (low-mid-high), and a `scale_color_gradientn()`, which creates an n-color gradient.

<p class = 'code'>Create 2-color gradient</p>
```{r colram, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradient(low = "#FCFC9A", high = "#F54927", 
                              aesthetics = c("fill", 'color'), 
                              name = "Temp. (F)") 

```
<br>
<p class = 'code'>Create diverging gradient</p>
For the divergent palette to be meaningful, you usually need to set the midpoint if it's not 0.
```{r colram2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "navy", mid = "#FCFC9A", high = "#F54927", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(lakes_temp$Value), 
                               name = "Temp. (F)") 

```
<br>
<p class = 'code'>Create diverging gradient with multiple colors</p>
Note the change in the legend by using `guide = 'legend'`. Default is `guide = 'colorbar'`. I also customized the breaks into 5-degree bins using `breaks()` and `seq()`.
```{r colram2b, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradientn(colors = c("#805A91", "#406AC2", "#FBFFAD", "#FFA34A", "#AB1F1F"), 
                               aesthetics = c("fill", 'color'),
                               guide = "legend",
                               breaks = c(seq(40, 85, 5)), 
                               name = "Temp. (F)") 

```

<div class="alert alert-info">
<strong>CHALLENGE: Create your own palette with at least three colors.</strong><br>
Hint: Start with p_heat to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal2d, fig.height = 4.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "#3E693D", mid = "#FDFFC7", high = "#7A6646", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(lakes_temp$Value), 
                               name = "Temp. (F)") 
```
</details>
</div>

</details>
<br>

