---
output: 
  html_document:
    css: custom_styles.css
---

```{r day2_gg, echo = F, include = F}
#------------------------------------
#    Day 2: Intro to ggplot Code 
#------------------------------------
```

```{r optchunk, echo = F}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align = 'center', fig.height = 4, fig.width = 6)
```

#### Intro to `ggplot2`
<details open><summary class='drop'>Intro to `ggplot2`</summary>
The `ggplot2` package is the most popular `R` package for plotting. It takes a little effort to learn how the pieces of a ggplot object fit together. However, once you get the hang of it, you can create and customize a large variety of attractive plots with just a few lines of `R` code. The package is called `ggplot2` because originally there was `ggplot`. The developer, Hadley Wickham, didn't want to break the original package to improve the package, so created ggplot2.  

[The ggplot2 online book]("https://ggplot2-book.org/") and [cheatsheets](https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf") can be very helpful while you are learning to use the `ggplot2` package.

The `ggplot2` package was developed using the grammar of graphics as the underlying philosophy, which basically breaks a plot up into individual building blocks related to aesthetics (e.g., color, size, shape), geometries (e.g. points, lines, boxes), and themes (e.g. axis label font size, legend placement, etc.).

Important concepts with `ggplot2`:
<ul>
<li><b>ggplot(data, aes()):</b> Every ggplot object starts with this line, which tells R the data you're plotting, and which variables you're plotting where in `aes()` argument.</li>
<li><b>data: </b> every plot requires data. The first argument for every `ggplot()`call is the data. This also means you can pipe data into a ggplot object. </li>
<li><b>aes: </b> Short for aesthetics. This is where you tell ggplot what your x and y variables are. If you want aesthetics, like color, fill, or size to vary by the data (e.g., color code a figure by park, use a dashed vs. solid line to distinguish between significant/non-significant trend, etc.), those variables are specified within the `aes()` argument either at the `ggplot()` level, or within the specific geom. </li>
<li><b>geom: </b> Short for geometries, geoms represent what you see in the plot, such as the points in a scatter plot, the lines in a trend plot, or the boxes in a boxplot.</li>
<li><b>scale: </b>Scales go hand in hand with `aes()`. If you specify `aes(color = park)`, then scale is where you can specify a custom color for each park instead of ggplot's default color scheme. The scale is where you can set the labels of groups in a legend (if different than how the data are labeled). You can also customize axis ranges, breaks, and labels with different scales. </li>
<li><b>theme: </b>This is where you can change the format of the figure, such as removing the gridlines in the default ggplot format, making axis labels larger, changing the position of the legend, etc.</li>
<li><b>facet: </b>Faceting the data allows you to graph multiple plots based on a grouping variable (e.g., site, species, year, etc.) in the same ggplot object.</li>
</ul>

</details>
<br>

#### Building a ggplot 
<details open><summary class = "drop">Building a ggplot object </summary>
We're going to use site-level photoplot % cover data from Ship Harbor to create the plot below, and will work through the code one piece at a time. 

```{r echo = F, fig.height = 5, fig.width = 7, fold.plot = F, fig.cap = "Mean percent cover by community type (panels) in Ship Harbor. Error bars represent +/- 1 SE."}
library(ggplot2) # load ggplot
library(dplyr)
pcov <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/SHIHAR_photoplot_cover.csv") # import data

cols <- c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
         "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
         "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")

shps <- c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
          "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25)

xrange <- range(pcov$Year)

ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  #geom_smooth(se = F, span = 0.75) +
  #geom_line(linewidth = 0.6) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```
<br>
<p class = 'code'>Import photoplot cover data from Ship Harbor</p>
```{r gg1, eval = F}
# load package
library(ggplot2) 
# import data
pcov <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/SHIHAR_photoplot_cover.csv") 
# check out the data
head(pcov) 
```
<br>
<p class = 'code'>Create the ggplot template of average cover over time.</p>
I'm going to assign this to an object named `p`, so we can build it one layer at a time. We're going to have unique colors and shapes for each level of CoverCode in our data, so we need to indicate that in the `aes()` along with our x and y variables. If we don't include color, fill, and shape in the `aes()`, the points would all be the same color (default = black) and shape (default is filled circle).
```{r gg2, fold.plot = F}
p <- ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                             color = CoverCode, 
                             fill = CoverCode,
                             shape = CoverCode))

p
```
</details><br>

<details open><summary class = "drop">Add point and error bars</summary>
<p class = 'code'>Add point and errorbar geometry</p>
The order of geometries is the order they're drawn. I prefer the look of the points after the error bars. 
```{r gg3, fold.plot = F}
p2 <- p + 
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover)) +
  geom_point()

p2
```
<br>
<p class = 'code'>Specify colors and shapes</p>
The default colors and symbols in ggplot aren't great. We're going to start by specifying our own colors and shapes manually. Then we'll use color palettes from different packages. Specifying shapes in R requires knowing the shape's symbol code. To view that, run `?points`, or search "pch in R plot" and you'll get the info below. Note that 0-14 are just lines with no fill. To change their color, use the color aesthetic. Symbols 15-20 are solid, but also use color to change their aesthetic. Symbols 21-25 have both a color (outline) and fill (inside) aesthetic.

<center><img src = "./images/r_pch_symbols.png" alt = 'R pch symbols' style = "width:250px;background:white;"><br>
<p class = "cap">Figure of symbol codes in R.</p></center>
<br>

```{r gg4, fold.plot = F}
p3 <- p2 + 
  scale_fill_manual(values = c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
                               "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
                               "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")) +
  scale_color_manual(values = c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
                               "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
                               "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")) +
  scale_shape_manual(values = c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
                                "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25))
p3
```
The code above was redundant and tedious to code. Changing colors means changing them in two places. Imagine you're making one of these plots for every rocky intertidal site in Acadia. It quickly becomes cumbersome. A more efficient way to code this is below. 
<br>
<p class = 'code'>Assign colors and shapes to variables </p>
Adding the name to the scales renames the legend title. But to keep the shapes and colors in the same legend, you have to name all of them the same name.
```{r gg4b, fold.plot = F}
cols <- c("ASCNOD" = "#C5B47B", "BARSPP" = "#A9A9A9", 
         "NONCOR" = "#574F91", "FUCSPP" = "#FFD560",
         "MUSSPP" = "#6F88BF", "REDGRP" = "#FF4C53")

shps <- c("ASCNOD" = 23, "BARSPP" = 24, "NONCOR" = 23,
          "FUCSPP" = 25, "MUSSPP" = 23, "REDGRP" = 25)

p3 <- p2 + 
  geom_point(color = 'dimgrey', size = 2.5) + # setting point outline to dark grey 
  scale_fill_manual(values = cols, name = "Species Group") +
  scale_color_manual(values = cols, name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group")
  
p3
```
</details><br>

<details open><summary class = 'drop'>Improve formatting</summary>
<p class = 'code'>Improve axis ticks and labels</p>
```{r gg5, fold.plot = F}
# Determine year range (so not hard coded/easily updated in future)
xrange <- range(pcov$Year)

p4 <- p3 + 
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") 
  
p4
```
<br>
<p class = 'code'>Improve theme components</p>
```{r gg6, fold.plot = F} 
p5 <- p4 +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), # angle x text
          panel.grid.major = element_blank(), # turns of major grids
          panel.grid.minor = element_blank(), # turns off minor grids
          panel.background = element_rect(fill = 'white', color = 'dimgrey'),# makes background white 
          legend.key = element_blank()) # removes square fill around symbols in legend

p5
```
<br>
<p class = 'code'>Facet on CommunityType</p>
```{r gg7, fold.plot = F, fig.height = 5, fig.width = 7,}
p6 <- p5 + facet_wrap(~CommunityType)

p6  
```
</details><br>

<details open><summary class = 'drop'>Full code and save figure</summary>
```{r gg8, fold.plot = F, fig.height = 5, fig.width = 7}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```
<br>
<p class = 'code'>Save plot to disk</p>
Copy and pasting figures from R into word documents or powerpoints usually results in a poor resolution figure. The better approach is to save figures to disk using `ggsave()`. The function saves the most recent figure that was generated in the Plots tab in the bottom right pane. You can also specify the ggplot object name, which would come before the file name. If you wanted a jpg or png instead of svg, just use that as the file extension type. The svg is a vectorize image, which for figures with lines, is usually the best quality image that won't become pixelated when zoomed. Only caveat is not all software supports svgs.
```{r ggs, eval = F}
ggsave("SHIHAR_photoplot_cover.svg", height = 8, width = 7)
```

</details>
<br>

<details open><summary class = 'drop'>Test your ggplot skills</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, make the point size 1.5.</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga1}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 1.5) + # changed this line
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, make the error bars wider.</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga2}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 1.2) + # changed this line
  geom_point(color = "dimgrey", size = 2.5) + 
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") + 
  scale_shape_manual(values = shps, name = "Species Group") + 
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Using the full plot code above, change the x axis label to "Year".</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga3}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) + 
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = "Year", y = "Avg Percent Cover") + # changed this line
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

</details>
</div>
</details><br>

#### Other geometries
<details open><summary class = 'drop'>Lines and smoothers</summary>
<p class = 'code'>Add line geometry to figure</p>
Note that for lines to plot properly in ggplot, you have to assign a grouping variable in the `aes()`. The fill, color, and shape aesthetics in the beginning already did that for us, but here's how it would look within `geom_line()` if you needed to do that. 
```{r gg11, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + geom_line(linewidth = 0.8, aes(group = CoverCode))
```
<br>
<p class = 'code'>Add smoother to figure</p>
The geom_smooth() plots a line assuming the y ~ x formula (unless you specify a different formula). By default the method is a LOESS smoother, but you can specify a range of methods, including linear regression by adding method = 'lm' to geom_smooth().

Note that I turned off the standard error ribbon that plots by default using se = FALSE. It’s too busy for this plot. I also don’t use the SE unless I’ve fit an actual model and checked the diagnostics. The status under the hood of geom_smooth() are also pretty black boxy, and I don’t always know if I can trust its calculation of SE.
```{r gg12, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + geom_smooth(se = F, span = 0.75)
```
<br>
<div class="alert alert-info">
<strong>CHALLENGE: How would you change the smoother in the code above from LOESS to linear model and make the line dashed? Hint: method = 'lm'.</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga3b}
p6 + geom_smooth(method = 'lm', se = F, linetype = 'dashed')
```
</details>
</div>
<br>

<p class = 'code'>Add horizontal dashed line called 50% line</p>
```{r gg13, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + geom_hline(aes(yintercept = 50), linetype = "dashed") 
```
<br>
<p class = 'code'>Add horizontal dashed line called 50% line and make it show in the legend</p>
```{r gg14, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + geom_hline(aes(yintercept = 50, linetype = "50% line")) +
     scale_linetype_manual(values = c("50% line" = "dashed"), 
                           name = "Threshold")
```
</details><br>

<details open><summary class = 'drop'>Theme changes</summary>
<p class = 'code'>Move legend to the bottom</p>
```{r gg15, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + theme(legend.position = 'bottom')
```

<div class="alert alert-info">
<strong>CHALLENGE: How would you make the legend title larger and bold, and legend text larger?</strong>
<details><summary class='drop2'>Answer</summary>
```{r gga3bc}
p6 + theme(legend.title = element_text(size = 12, face = 'bold'),
           legend.text = element_text(size = 11))
```
</details>
</div>
<br>

<p class = 'code'>Change facet strips fill, color, and font size</p>
```{r gg16, fig.height = 5, fig.width = 7, fold.plot = F}
p6 + theme(strip.background = element_rect(fill = "#F5F0DC", color = "black"),
           strip.text = element_text(size = 10))
```
</details><br>

<details open><summary class = 'drop'>Other geometries</summary>
<p class = 'code'>Stacked bars instead of points </p>
```{r gg17, fig.height = 5, fig.width = 7, fold.plot = F}
ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_bar(stat = 'identity', position = 'fill', color = 'dimgrey') +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Median. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), # turns of major grids
        panel.grid.minor = element_blank(), # turns off minor grids
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)

```

<p class = 'code'>Bars with error bars instead of points (note I filtered on Barnacle CommunityType)</p>
```{r gg18, fig.height = 5, fig.width = 7, fold.plot = F}
ggplot(data = pcov |> filter(CommunityType == "Barnacle"), # note filter 
       aes(x = Year, y = avg_cover, 
           color = CoverCode, group = CoverCode,
           fill = CoverCode, shape = CoverCode)) +
  geom_bar(stat = 'identity', position = 'dodge', color = 'dimgrey') + # new line
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), linewidth = 0.6) +
  scale_fill_manual(values = cols, aesthetics = c("fill", "color"), 
                    name = "Species Group") +
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = NULL, y = "Avg. Percent Cover") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CoverCode) # Different facet


```
</details>
<br>

#### Multipanels {.tabset}

<details open><summary class = 'drop'>Multi-panel plots via patchwork</summary>
Faceting is helpful when your observations are all within the same column. But say you have data in multiple columns and want to arrange those plots into a grid. Faceting won't help because the data to plot are in different columns. There are multiple packages that make it easy to arrange multiple plots into a grid to look similar to faceted plots. Packages include `grid` (and `gridExtra`), `cowplot`, `ggpubr`, and `patchwork`. We're going to use `patchwork`, a relative newcomer, and one of the easiest I've found to code and customize. Here we're going to plot pH, temperature, DO, and conductance for Jordan Pond in Acadia NP and arrange them using patchwork.

The patchwork package has a lot of options to customize plot layouts. See the <a href="https://patchwork.data-imaginist.com/">patchwork</a> package website for more information.

<p class = 'code'>Load patchwork and read in water chemistry data</p>
```{r pw}
# load packages
library(ggplot2) 
library(patchwork) # multipanel plots

# load data
chem <- read.csv(
  "https://raw.githubusercontent.com/KateMMiller/MMA_R_Training_2026/refs/heads/main/data/ACAD_Jordan_Pond_water_chem.csv")

# make date field a date
chem$date <- as.Date(chem$date, format = "%Y-%m-%d")
```

<p class = 'code'>Create a ggplot object for each parameter</p>
```{r pw2}
# pH plot
p_pH <-
  ggplot(chem, aes(x = date, y = pH)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "pH", x = "Year") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# temp plot
p_temp <-
  ggplot(chem, aes(x = date, y = Temp_F)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "Temp (F)", x = "Year") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
 
# Diss. Oxygen plot
p_do <-
  ggplot(chem, aes(x = date, y = DO_mgL)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "DO (mg/L)", x = "Year") +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# Conductance plot
p_cond <-
  ggplot(chem, aes(x = date, y = SpCond_uScm)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "dimgrey", alpha = 0.5, size = 2) +
  labs(y = "Spec. Cond. (uScm)", x = "Year")+
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
<br>

<p class='code'>Arrange plots using patchwork</p>
This is almost too easy to be true, but it really is this easy with patchwork. The patchwork package includes a bunch of options to customize sizes, add annotation, sharing axes across plots. 
```{r patch1, fold.plot = F, fig.height = 4, fig.width = 5}
p_pH + p_temp + p_do + p_cond
```

<p class='code'>Arrange plots using patchwork in column of 4 and share x axis. </p>
You can also collect the legend using a similar approach to collecting the axes.
```{r patch2, fold.plot = F, fig.height = 6, fig.width = 3}
library(patchwork)
p_pH / p_temp / p_do / p_cond + plot_layout(axes = "collect_x")
```

</details>
<br>




