---
output: 
  html_document:
    css: custom_styles.css
---

#### Palettes
<details open><summary class = 'drop'>External ggplot Palettes</summary>
A number of R packages have color palettes available that are colorblind friendly. The two most commonly used are RColorBrewer and viridis. Both packages include palettes that are for three main types of data: 
 <ul>
 <li><b>Sequential</b>: for continuous variables that grade from low to high or vice versa. These tend to be one hue that increases in saturation as values increase.</li>
 <li><b>Diverging</b>: for data where low values have different colors than high values (e.g. starts red, then grades to blue).</li>
 <li><b>Qualitative</b>: data for categorical variables where all colors are a similar saturation but different hues.</li>
 </ul>

<details open><summary class='drop2'>RColorBrewer palette</summary>
The palettes in RColorBrewer can be viewed by running the code below. The first group shows the sequential palettes (e.g. YlOrRd - Yellow Orange Red). The second group shows the qualitative colors. The last group shows the diverging palettes. The main drawback of these palettes is they are limited by the number of levels in your data. So, if you specify `Set2` to color code different levels of a factor, there are only 8 colors available to you. If your factor has more than 8 levels (e.g., 9 sites, 10 parks, etc.), then the levels beyond 8 won't get plotted and you'll get a warning in the console similar to what we saw for ggplot's default number of symbols.

```{r day3_gp, echo = F, include = F}
#------------------------------------
#     Day 3: ggplot Palettes Code 
#------------------------------------
```

<p class = 'code'>View RColorBrewer palettes</p>
```{r rbcolpal_code, eval = F}
display.brewer.all(colorblindFriendly = TRUE)
```
<center><img src = "./images/RColorBrewer_palettes.png" alt = 'RColorBrewer palettes' style = "width:350px;"></center>
<br>

Going back to the photoplot cover plots we made before, we'll use RColorBrewer to color code each CoverCode instead of doing this manually. We'll build the plot in the next chunk, that we then change the color palettes with in later plots.

<br>
<p class = 'code'>Create basic plot</p>
```{r tempp2}
p_pal <- ggplot(data = pcov, aes(x = Year, y = avg_cover, 
                        color = CoverCode, group = CoverCode,
                        fill = CoverCode, shape = CoverCode)) +
  geom_errorbar(aes(ymin = avg_cover - se_cover, ymax = avg_cover + se_cover), 
                linewidth = 0.6) +
  geom_point(color = "dimgrey", size = 2.5) + 
  scale_shape_manual(values = shps, name = "Species Group") +
  scale_x_continuous(limits = c(xrange[1] - 1, xrange[2] + 1),
                     breaks = c(seq(xrange[1] - 1, xrange[2] + 1, 2))) +
  labs(x = "Year", y = "Avg Percent Cover") + # changed this line
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = 'white', color = 'dimgrey'), 
        legend.key = element_blank()) +
  facet_wrap(~CommunityType)
```

<br>

<p class = 'code'>Use Set2 palette for species groups</p>
```{r set2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Species Group", palette = "Set2", 
                            aesthetics = c("fill", "color")) 

```

Note how I used the aesthetics in the `scale_color_brewer()` to set fill and color as the same time. We could have done this in the code above too.

<br>

<p class = 'code'>Use Dark2 palette on temperature plot</p>
```{r colbred2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Species Group", palette = "Dark2", 
                            aesthetics = c("fill", "color")) 

```
</details>
<br>

<div class="alert alert-info">
<strong>CHALLENGE: How would you specify the 'RdYlBu' palette instead of the ones used above?</strong><br>
Hint: Start with p_pal to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal, fig.height = 4.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Species Group", palette = "RdYlBu", 
                            aesthetics = c("fill", "color"))  
```
</details>
</div>

<details open><summary class = 'drop2'>viridis palettes</summary>
The viridis package comes with 8 palettes. The benefit of viridis is the number of levels is not limited to 8 like RColorBrewer. The palette options are below for 12 levels. 

<center><img src = "./images/viridis/viridis-scales.png" alt = 'viridis palettes' style = "width:450px;"></center>

<br>

<p class = 'code'>View viridis palettes with hexcodes</p>
You can view the hexcodes of the different palettes by running the code below. Just change `viridis()` to one of the other palette names to get the hexcodes for those levels. 
```{r vir, eval = F}
# viridis 
scales::show_col(viridis(12), cex_label = 0.45, ncol = 6)
```
<center><img src = "./images/viridis/viridis.png" alt = 'viridis palette' style = "width:350px;"></center>
<br>
<p class = 'code'>Use viridis default palette on CoverCode</p>
The `scale_color_viridis_d()` selects the viridis palette option (purple, green, yellow) for discrete values (i.e. categories). For a continuous scale (e.g. temperature), you would specify `scale_color_viridis_c()`. 
```{r vir2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal + scale_color_viridis_d(name = "Species Group", aesthetics = c("fill", "color"))  #default viridis 
```

<p class = 'code'>Use turbo palette on CoverCode</p>
The `scale_color_viridis_d()` selects the viridis palette option (purple, green, yellow) for discrete values (i.e. categories). For a continuous scale (e.g. temperature), you would specify `scale_color_viridis_c()`. 
```{r vir2b, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_pal + scale_color_viridis_d(name = "Species Group", aesthetics = c("fill", "color"), option = 'turbo') 
```
</details>
<br>
<details open><summary class = 'drop2'>Continuous palette with heatmaps</summary>
Heatmaps via `geom_tile()` are a place where viridis palettes are especially helpful producing useful sequential or diverging color palettes. We'll use the temperature data to plot heatmaps by month for each site. Heatmaps are a bit different than other plots we've seen, as the x and y values create a discrete grid, and the color in the cell represents the value for that level of x and y. That means we have to change how the x, y and color aesthetics are specified. Here we will plot temperature by month and year faceted on site.

<p class = 'code'>Basic heatmap code </p>
Note the use of base R's `month.abb` to set the labels on the x-axis. The `month.abb` is a vector of the 12 months abbreviated as 3 letters. By setting 5:10, I'm taking the months May - Oct.
```{r virheat}
p_heat <- 
ggplot(chem, aes(x = mon, y = year, color = Temp_F, fill = Temp_F)) + 
  theme_bw() +
  geom_tile() + 
  labs(y = "Year", x = "Month") +
  scale_x_continuous(breaks = c(5, 6, 7, 8, 9, 10),
                     limits = c(4, 11), 
                     labels = month.abb[5:10]) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```

<p class = 'code'>Plot heatmap with viridis continuous palette </p>
```{r virheat1, fold.output = F, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_viridis_c(name = "Temp. (F)", aesthetics = c("fill", "color")) 
```

<p class = 'code'>Plot heatmap with plasma continuous palette, reverse scale </p>
```{r virheatp, fold.output = F, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_viridis_c(name = "Temp. (F)", aesthetics = c("fill", "color"), 
                               option = "plasma", direction = -1) 
```

</details>
</details>
<br>
<details open><summary class = 'drop'>Create your own color ramp</summary>
You can also create your own color ramp via `scale_color_gradient()`, which creates a 2-color gradient,  `scale_color_gradient2()`, which creates a diverging color gradient (low-mid-high), and a `scale_color_gradientn()`, which creates an n-color gradient.

<p class = 'code'>Create 2-color gradient</p>
```{r colram, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradient(low = "#FCFC9A", high = "#F54927", 
                              aesthetics = c("fill", 'color'), 
                              name = "Temp. (F)") 

```
<br>
<p class = 'code'>Create diverging gradient</p>
For the divergent palette to be meaningful, you usually need to set the midpoint if it's not 0.
```{r colram2, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "navy", mid = "#FCFC9A", high = "#F54927", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(chem$Temp_F), 
                               name = "Temp. (F)") 

```
<br>
<p class = 'code'>Create diverging gradient with multiple colors</p>
Note the change in the legend by using `guide = 'legend'`. Default is `guide = 'colorbar'`. I also customized the breaks into 5-degree bins using `breaks()` and `seq()`.
```{r colram2b, fold.plot = F, fig.height = 3.5, fig.width = 7}
p_heat + scale_color_gradientn(colors = c("#805A91", "#406AC2", "#FBFFAD", "#FFA34A", "#AB1F1F"), 
                               aesthetics = c("fill", 'color'),
                               guide = "legend",
                               breaks = c(seq(40, 85, 5)), 
                               name = "Temp. (F)") 

```

<div class="alert alert-info">
<strong>CHALLENGE: Create your own palette with at least three colors.</strong><br>
Hint: Start with p_heat to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal2d, fig.height = 4.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "#3E693D", mid = "#FDFFC7", high = "#7A6646", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(chem$Temp_F), 
                               name = "Temp. (F)") 
```
</details>
</div>

</details>
<br>

#### CTD data
<details open><summary class='drop'>Example code for plotting CTD data in ggplot. </summary>
<p class = 'code'>Load libraries and import data </p>
```{r ctd1}
library(tidyverse)
library(readxl)

ctd_mma <- read_xlsx("./data/PR_PF_2903444 (2).xlsx") |> data.frame()
```

<p class = 'code'>Graph data with reversed Y axis and color coded by station</p>
```{r fold.plot = F, fig.height= 8, fig.width = 6}
ggplot(ctd_mma, aes(x = `TEMP..degree_Celsius.`, 
                    y = `PRES..decibar.`,
                    group = Station, 
                    color = Station)) +
  geom_line() + 
  theme_bw() +
  labs(x = "Temp. (C)", y = "Pressure (dbars)") +
  scale_color_gradientn(colors = c("#805A91", "#406AC2", "#FBFFAD", "#FFA34A", "#AB1F1F"), 
                        aesthetics = c('color'),
                        guide = "legend", # makes legend distinct, rather than color band
                        breaks = 1:15, # number of stations
                        name = "Station ID") +
  scale_y_reverse() + # flip y axis
  scale_x_continuous(limits = c(0, 30),
                     breaks = seq(0, 30, 5),
                     position = 'top') + # plot x-axis on top
  theme(legend.position = 'bottom')

```
</details>
<br>
