#### Day 3 Questions
<details open><summary class='drop'>Load Data and Packages</summary>
<br>

```{r day3_ch, echo = F, include = F}
#------------------------------------
#       Day 3: Challenges Code 
#------------------------------------
```

<p class = 'code'>Load packages and prep data for ggplot sections</p>
```{r loadwdat}
# Water chemistry data for ggplot section
library(dplyr)
library(ggplot2)
library(patchwork) # for arranging ggplot objects
library(RColorBrewer) # for palettes
library(viridis) # for palettes
chem <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_water_chemistry_data.csv")

chem <- chem |> mutate(date = as.Date(EventDate, "%m/%d/%Y"),
                       year = as.numeric(format(date, "%Y")),
                       mon = as.numeric(format(date, "%m")),
                       doy = as.numeric(format(date, "%j"))) 

ACAD_lakes <- c("ACBUBL", "ACEAGL", "ACECHO", "ACJORD", 
                "ACLONG", "ACSEAL", "ACUHAD", "ACWHOL")

lakes_temp <- chem |> filter(SiteCode %in% ACAD_lakes) |> 
  filter(Parameter %in% "Temp_F") 
```
</details>
</details><br>

<details open><summary class='drop'>Customizing ggplot</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Review labels from Day 2. How would you change the X-axis label to "Year", and the Y-axis label to "Temp. (F)"?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3labd3}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point(color = "dimgrey", size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(x = "Year", y = "Temp. (F)") # answer
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the smooth from LOESS to linear?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3lmc1}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_point(color = "dimgrey", alpha = 0.5) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake")  +
  labs(x = "Year", y = "Temp. (F)") 
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Recreate the plot below</strong>
Note that the symbol outline is black. The alpha level is 0.6.
```{r challd3copyc1, echo = F, fold.plot = F, fig.height = 7, fig.width = 5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')
```

<details><summary class='drop2'>Answer</summary>
```{r challd3copyac1, fig.height = 7, fig.width = 5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')

```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the plot called p_site2 so that x-axis labels were every 4 years, and angle is 45 degrees instead of 90?</strong><br>
Hint: start with p_site, so you don't have to write that much code. You'll also need to tweak vjust and hjust to make the labels line up properly.
<details><summary class='drop2'>Answer</summary>
```{r challyearsc1, fig.height = 4.5, fig.width = 7}
p_site + 
  scale_x_date(date_labels = "%Y", date_breaks = "4 years") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
</details>
</div>
</details>
<br>

<details open><summary class = 'drop'>ggplot Palettes</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you specify the 'RdYlBu' palette instead of the ones used above?</strong><br>
Hint: Start with p_pal to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpalc1, fig.height = 4.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Lake", palette = "RdYlBu", aesthetics = c("fill", "color"))  +
         guides(fill = guide_legend(override.aes = list(alpha = 1))) # makes symbols not transparent in legend
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Create your own palette with at least three colors.</strong><br>
Hint: Start with p_heat to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal2c, fig.height = 4.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "#3E693D", mid = "#FDFFC7", high = "#7A6646", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(lakes_temp$Value), 
                               name = "Temp. (F)") 
```
</details>
</div>

