#### Day 3 Questions
<details open><summary class='drop'>Load Data and Packages</summary>
<br>

```{r day3_ch, echo = F, include = F}
#------------------------------------
#       Day 3: Challenges Code 
#------------------------------------
```

<details open><summary class = 'drop2'>Pivoting tables</summary>
<p class = 'code'>Load data and dplyr</p>
```{r loaddat3}
library(dplyr)

# bat capture data
bat_cap <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/example_bat_capture_data.csv")
```
</details>
<br>
<details open><summary class = 'drop2'>Joining tables</summary>
<p class = 'code'>Load data and dplyr</p>
```{r loaddat3j}
library(dplyr)
# tree data
trees <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_data.csv")
# tree species table
spp_tbl <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_tree_species_table.csv")
```
</details>
<br>
<details open><summary class = 'drop2'>Dates and times</summary>

<p class = 'code'>Load and prep data.</p>
```{r loadhobodat}
# Hobo Temp data
temp_data <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/HOBO_temp_example.csv", skip = 1)[,1:3]
colnames(temp_data) <- c("index", "temp_time", "tempF")
temp_data$timestamp_temp <- as.POSIXct(temp_data$temp_time, 
                                       format = "%m/%d/%Y %H:%M", 
                                       tz = "America/New_York")
```
</details>
<br>
<details open><summary class = 'drop2'>ggplot sections</summary>
<p class = 'code'>Load packages and prep data for ggplot sections</p>
```{r loadwdat}
# Water chemistry data for ggplot section
library(dplyr)
library(ggplot2)
library(patchwork) # for arranging ggplot objects
library(RColorBrewer) # for palettes
library(viridis) # for palettes
chem <- read.csv("https://raw.githubusercontent.com/KateMMiller/IMD_R_Training_2026/refs/heads/main/data/NETN_water_chemistry_data.csv")

chem <- chem |> mutate(date = as.Date(EventDate, "%m/%d/%Y"),
                       year = as.numeric(format(date, "%Y")),
                       mon = as.numeric(format(date, "%m")),
                       doy = as.numeric(format(date, "%j"))) 

ACAD_lakes <- c("ACBUBL", "ACEAGL", "ACECHO", "ACJORD", 
                "ACLONG", "ACSEAL", "ACUHAD", "ACWHOL")

lakes_temp <- chem |> filter(SiteCode %in% ACAD_lakes) |> 
  filter(Parameter %in% "Temp_F") 
```
</details>
</details><br>

<details open><summary class='drop'>Pivoting tables</summary>
<div class="alert alert-info">
<strong>Pivot the `bat_sum` data frame on year instead of species, so that you have a column for every year of captures. Remember to avoid column names starting with a number. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c1, fold.output = F}
bat_wide_yr <- pivot_wider(bat_sum, names_from = Year, 
                           values_from = num_indiv, 
                           values_fill = 0, 
                           names_prefix = "yr")
head(bat_wide_yr)
```
</details>
</div>

<div class="alert alert-info">
<strong>Pivot the resulting data frame from the previous question to long on the years columns, and remove the "yr" from the year names using `names_prefix = 'yr'`. </strong>

<details><summary class='code2'>Answer</summary>
```{r pivwide2c2, fold.output = F}
bat_long_yr <- pivot_longer(bat_wide_yr, 
                            cols = -c(Site, sppcode),
                            names_to = "Year", 
                            values_to = "num_indiv", 
                            names_prefix = "yr") # drops this string from values
```
</details>
</div>
</details><br> 

<details open><summary class='drop'>Joining Tables</summary>
<div class="alert alert-info">
<strong>Join the NETN_tree_data.csv and NETN_tree_species_table.csv to connect the common name to the tree data.</strong>

<details><summary class='drop2'>Answer</summary>
```{r challtr2bc1}
# find the columns in common
intersect(names(spp_tbl), names(trees)) # TSN and ScientificName
# left join species to trees, because don't want to include species not found in tree data
trees_spp <- left_join(trees, 
                       spp_tbl |> select(TSN, ScientificName, CommonName), 
                       by = c("TSN", "ScientificName"))

head(trees_spp)
```
</details>
</div>

<div class="alert alert-info">
<strong>Find a species in the NETN tree data that doesn't have a match in the species table.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challtr2d}
# find the columns in common
intersect(names(spp_tbl), names(trees)) # TSN and ScientificName
# anti join of trees against species table, selecting only columns of interest
anti_join(trees, spp_tbl, by = c("TSN", "ScientificName")) |> 
  select(ParkUnit, PlotCode, SampleYear, ScientificName)
```
</details>
</div>

</details>
</br>

<details open><summary class='drop'>Dates and Times</summary>
<div class="alert alert-info">
<strong>Question: How would you return date1 as YYYYMMDD (20260312)?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald31dc}
format(date1, format = "%Y%m%d")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you create a list of dates in 2026 that are evenly spaced by 3 months?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32bc}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "3 months")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you create a list of dates in 2026 that are evenly spaced by 1 week?</strong>
<details><summary class='drop2'>Answer</summary>
```{r chald32bd}
date_list <- as.Date(c("01/01/2026", "12/31/2026"), format = "%m/%d/%Y") 
seq.Date(date_list[1], date_list[2], by = "1 week")
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you extract the month as a number ranging from 1-12 in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatec1}
temp_data$month_num <- as.numeric(format(temp_data$timestamp_temp, "%m"))
head(temp_data)
```
</details>
</div>

<div class="alert alert-info">
<strong>How would you extract the julian date in temp_data?.</strong>
<details><summary class='drop2'>Answer</summary>
```{r challdatejc1}
temp_data$julian <- as.numeric(format(temp_data$timestamp_temp, "%j"))
head(temp_data)
```
</details>
</div>

</details>
<br>

<details open><summary class='drop'>Customizing ggplot</summary>
<div class="alert alert-info">
<strong>CHALLENGE: Review labels from Day 2. How would you change the X-axis label to "Year", and the Y-axis label to "Temp. (F)"?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3labd3}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_point(color = "dimgrey", size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(x = "Year", y = "Temp. (F)") # answer
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the smooth from LOESS to linear?</strong>
<details><summary class='drop2'>Answer</summary>
```{r challd3lmc1}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_point(color = "dimgrey", alpha = 0.5) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake")  +
  labs(x = "Year", y = "Temp. (F)") 
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Recreate the plot below</strong>
Note that the symbol outline is black. The alpha level is 0.6.
```{r challd3copyc1, echo = F, fold.plot = F, fig.height = 7, fig.width = 5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')
```

<details><summary class='drop2'>Answer</summary>
```{r challd3copyac1, fig.height = 7, fig.width = 5, fig.align = 'center'}
ggplot(lakes_temp, aes(x = date, y = Value, color = SiteCode, 
                       fill = SiteCode, shape = SiteCode)) + 
  theme_bw() +
  geom_smooth(se = F, span = 0.5) +
  geom_point(color = "black", alpha = 0.6, size = 2) +
  scale_color_manual(values = site_cols, name = "Lake") +
  scale_fill_manual(values = site_cols, name = "Lake") +
  scale_shape_manual(values = site_shps, name = "Lake") +
  labs(y = "Temp. (F)", x = "Year") +
  facet_wrap(~SiteCode, ncol = 2) +
  theme(legend.position = 'bottom')

```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: How would you change the plot called p_site2 so that x-axis labels were every 4 years, and angle is 45 degrees instead of 90?</strong><br>
Hint: start with p_site, so you don't have to write that much code. You'll also need to tweak vjust and hjust to make the labels line up properly.
<details><summary class='drop2'>Answer</summary>
```{r challyearsc1, fig.height = 4.5, fig.width = 7}
p_site + 
  scale_x_date(date_labels = "%Y", date_breaks = "4 years") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
</details>
</div>
</details>
<br>

<details open><summary class = 'drop'>ggplot Palettes</summary>
<div class="alert alert-info">
<strong>CHALLENGE: How would you specify the 'RdYlBu' palette instead of the ones used above?</strong><br>
Hint: Start with p_pal to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpalc1, fig.height = 4.5, fig.width = 7}
p_pal +  scale_color_brewer(name = "Lake", palette = "RdYlBu", aesthetics = c("fill", "color"))  +
         guides(fill = guide_legend(override.aes = list(alpha = 1))) # makes symbols not transparent in legend
```
</details>
</div>

<div class="alert alert-info">
<strong>CHALLENGE: Create your own palette with at least three colors.</strong><br>
Hint: Start with p_heat to save time coding.
<details><summary class='drop2'>Answer</summary>
```{r challpal2c, fig.height = 4.5, fig.width = 7}
p_heat + scale_color_gradient2(low = "#3E693D", mid = "#FDFFC7", high = "#7A6646", 
                               aesthetics = c("fill", 'color'),
                               midpoint = mean(lakes_temp$Value), 
                               name = "Temp. (F)") 
```
</details>
</div>

</details>
<br>