
<details open><summary class = 'drop'>Help Documentation</summary>
There are a number of options to get help with R. If you're trying to figure out how to use a function, you can type ?function_name. For example `?plot` will show the R documentation for that function in the Help panel.

<p class = 'code'>Get help for the functions below</p>
```{r, c4, echo=T, eval=F}
?plot
?dplyr::filter
```

You can also press F1 while the cursor is on a function name to access the help for that function. Help documents in R are standardized to help you find what you're looking for. 
<ul>
<li>Top left shows the function with the {package}. Base means it's a function in the base R install.</li>
<li><b>Description:</b> tell you what the function is. </li>
<li><b>Usage:</b> tells you what the arguments are. The "..." means there are other potential arguments, but isn't something we need to talk about right now. </li>
<li><b>Arguments:</b> define the arguments and what their inputs take. For example, if an argument is TRUE or FALSE, or a text string. </li>
<li><b>Value</b>: Describes more about the function (not always included) </li>
<li><b>See Also: </b>Sometimes functions build on other functions. This section links to similar or building block functions. </li>
<li><b>Examples: </b>Functions that provide good examples are invaluable. Sometimes these are afterthoughts or not included in help documentation, which is too bad. Unfortunately, base R functions tend to have some of the most obscure, hard to understand examples.</li>
</ul>
</details>
<br>

<details open><summary class = 'drop'>Troubleshooting errors</summary>
Great online resources to find answers to questions include <a href="https://stackexchange.com/">Stackexchange</a>, and <a href="https://stackoverflow.com//">Stackoverflow</a>. Google searches are usually my first step, and I include "in R" and the package name (if applicable) in every search related to R code. If you're troubleshooting an error message, copying and pasting the error message verbatim into a search engine often helps. 

Don't hesitate to reach out to colleagues for help as well! If you are stuck on something and the answers on Google are more confusing than helpful, don't be afraid to ask a human. Every experienced R programmer was a beginner once, so chances are they've encountered the same problem as you at some point. There is an R-focused <a href = "https://teams.microsoft.com/l/channel/19%3A114a351486924a7d92eca92dfe5235dd%40thread.tacv2/%5B%20%20%20%20Data%20Sci%20Community%20of%20Practice%20%5D?groupId=066491e4-d2ab-4a8a-931b-c96b970d24a8&tenantId=0693b5ba-4b18-4d7b-9341-f32f400a5494"> Data Science Community of Practice </a> for I&M folks, which anyone working in R (regardless of experience!) is invited and encouraged to join. 

</details>
<br>

<details open><summary class = 'drop'>Common errors and how to fix them</summary>
<ol>
<li><p class = "code">Unmatched parenthesis</p></li>
```{r par, eval = F}
mean_x <- mean(c(1, 3, 5, 7, 8, 21) # missing closing parentheses
mean_x <- mean(c(1, 3, 5, 7, 8, 21)) # correct
               
```

<li><p class = "code">Unmatched quotes</p></li>
```{r birb, eval = F}
birds <- c("black-capped chickadee", "golden-crowned kinglet, "wood thrush") # missing quote after kinglet
```
```{r birb2, eval = F}
birds <- c("black-capped chickadee", "golden-crowned kinglet", "wood thrush") # corrected

```

<li><p class = "code">Missing a comma between elements</p></li>
```{r birb3, eval = F}
birds <- c("black-capped chickadee", "golden-crowned kinglet" "wood thrush") # missing comma after kinglet
birds <- c("black-capped chickadee", "golden-crowned kinglet", "wood thrush") # corrected
```

<li><p class = "code">Misspelled function name</p></li>
```{r typo, eval = F}
x_mean <- maen(x) # misspelled mean
x_mean <- mean(x) # Corrected

```

<li><p class = 'code'>Incorrect use of dimensions with brackets</p></li>
```{r brac, error = T}
# Missing comma to indicate subsetting rows (records)
ACAD_wetland2 <- ACAD_wetland[!is.na(ACAD_wetland$Site_Name)]
# Correct
ACAD_wetland2 <- ACAD_wetland[!is.na(ACAD_wetland$Site_Name), ]

```
</ol>

Other resources that may help: 
<ul>
<li><a href = "https://www.r-bloggers.com/2023/02/top-10-errors-in-r-and-how-to-fix-them/">Top 10 errors in R and how to fix them blog</a></li>
<li><a href = "https://bioinformatics.ccr.cancer.gov/docs/btep-coding-club/CC2023/Troubleshooting_Rprog/">Troubleshooting R</a></li>
<li><a href = "https://r4ds.hadley.nz/workflow-help.html">R for Data Science: Chapter 8 Getting Help</a></li>
</ul>
<br>
<br>
